---
title: "Poisson GLM"
author: "Raphael Spagolla"
date: "2026 03 04"
output:
  html_document:
    df_print: paged
  pdf_document:
    latex_engine: xelatex
    number_sections: true
fontsize: 11pt
geometry: margin=2.5cm
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = FALSE
)
```
```{r}
# 1) Paths
path_rds <- "data_processed/heapo/heapo_modelling.rds"

# 2) Load
dat <- readRDS(path_rds)

# 3) Quick check
str(dat)
dim(dat)
names(dat)
head(dat)
```

### Interpretation of the Count Response Variable

The constructed response variable `n_high_days` represents the **number of days with high electricity consumption within a given month for each household**.

More precisely:
- `high_consumption = 1` if daily consumption exceeds the median
- For each household–month, these are summed to obtain `n_high_days`

Thus, the model explains the **frequency of high-consumption days per month**.

An offset `log(n_days)` is included to account for different month lengths, so the model estimates a **daily rate**.

```{r}
# 4) Create period (month)
dat <- dat %>%
  mutate(year_month = format(date, "%Y-%m"))

# 5) Aggregate to count per household-month
dat_count <- dat %>%
  group_by(Household_ID, year_month) %>%
  summarise(
    n_high_days = sum(high_consumption, na.rm = TRUE),
    n_days = n(),
    # optional: average predictors over period
    heating_degree_days = mean(heating_degree_days, na.rm = TRUE),
    temp_avg = mean(temp_avg, na.rm = TRUE),
    living_area = first(living_area),
    n_residents = first(n_residents),
    building_type = first(building_type),
    heatpump_type = first(heatpump_type),
    .groups = "drop"
  )

# 6) Check
head(dat_count)
summary(dat_count$n_high_days)
```









