---
# title: "Linear Model"
# author: "Tharrmeehan Krishnathasan"
# date: "2026 04 09"
output:
  pdf_document:
    latex_engine: xelatex
    toc: true
    toc_depth: '3'
  html_document:
    toc: true
    toc_float: true
    toc_depth: 3
    number_sections: true
    theme: flatly
    code_folding: hide
subtitle: Applied Machine Learning and Predictive Modelling 1 – FS 2026
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = FALSE
)
library(ggplot2)
library(dplyr)
library(broom)
library(car)
library(knitr)
set.seed(42)
```
```{r import-lm, echo=FALSE, message=FALSE, warning=FALSE}
heapo_lm   <- readRDS("../models/linear_model/heapo_lm.rds")
heapo_data <- readRDS("../data_processed/heapo/heapo_modelling.rds")

# Refit the simple model for comparison
heapo_mlm_simple <- lm(
  log_kWh_total ~ heating_degree_days +
    sunshine_hours +
    humidity_avg +
    living_area +
    n_residents +
    building_type +
    heatpump_type,
  data = heapo_data
)
```

# Linear Model

## Objective

The aim of this analysis is to model daily household energy consumption ($\text{log\_kWh\_total}$), the natural logarithm of total daily kilowatt-hour usage. Log-transforming energy consumption is standard practice: it is strictly positive, right-skewed on the original scale, and approximately normal after transformation, making linear regression appropriate.

Predictors include weather conditions (heating degree days, sunshine hours, humidity), building characteristics (building type, living area, heat pump type), and household composition (number of residents). By including interaction terms, the model also captures how the sensitivity to cold weather differs across building and heat pump types.

## Model Setup

Two models are estimated:

**Simple MLR** — all seven predictors with main effects only. This serves as an interpretable baseline capturing the additive relationships between predictors and log energy consumption:

$$
\log(\text{kWh}_i) = \beta_0 + \beta_1\,\text{HDD}_i + \beta_2\,\text{Sunshine}_i + \cdots + \beta_7\,\text{HeatpumpType}_i + \varepsilon_i
$$

**Extended MLR** — adds two domain-motivated interaction terms:

- $\text{heating\_degree\_days} \times \text{heatpump\_type}$: different heat pump technologies have different efficiencies in cold conditions, so the effect of cold on consumption should vary by pump type.
- $\text{heating\_degree\_days} \times \text{building\_type}$: well-insulated houses are less sensitive to outdoor temperature than poorly-insulated ones.

All coefficients are on the log scale: a coefficient of 0.05 means a 5% multiplicative change in kWh for a one-unit increase in the predictor.

Note that the panel structure of the data (multiple daily observations per household) means the i.i.d. assumption of standard OLS is not fully met — standard errors may be slightly understated due to within-household correlation.

## Model Results

```{r lm-coef-table, echo=FALSE}
tidy(heapo_lm) %>%
  filter(term != "(Intercept)") %>%
  mutate(
    estimate  = round(estimate, 4),
    std.error = round(std.error, 4),
    statistic = round(statistic, 2),
    p.value   = signif(p.value, 3)
  ) %>%
  knitr::kable(
    col.names = c("Predictor", "Estimate", "Std. Error", "t value", "p value"),
    caption   = "Extended MLR: coefficient estimates"
  )
```

All predictors are statistically significant at the 0.001 level, with the exception of `humidity_avg` (p = 0.165), which does not contribute meaningfully once the other variables are included. Key effects on log energy consumption, holding all other variables constant:

**Heating degree days** (β = 0.021, p < 0.001): Each additional HDD increases log consumption by ~2.1% for air-source pump households in flats — the baseline group. This is the reference effect before interaction adjustments.

**Sunshine hours** (β = −0.015, p < 0.001): An additional sunshine hour reduces consumption by ~1.5%, reflecting lower heating demand on brighter days.

**Number of residents** (β = 0.134, p < 0.001): Each additional resident increases consumption by ~14.3% ($e^{0.134} - 1$), the second strongest effect after the temperature-building interaction.

**Building type — house** (β = 0.220, p < 0.001): Houses consume ~24.6% more than flats at the baseline temperature, even before accounting for their greater sensitivity to cold (see interaction below).

**Heat pump type — ground-source** (β = −0.143, p < 0.001): Ground-source pump households consume ~13.3% less than air-source households at the baseline temperature.

**Interaction: HDD × house** (β = 0.039, p < 0.001): Houses are substantially *more* sensitive to cold than flats — each additional HDD increases log consumption by an additional 3.9% for house households. This is the largest interaction effect and confirms that houses are less thermally efficient.

**Interaction: HDD × ground-source** (β = −0.003, p < 0.001): Ground-source pump households are slightly *less* sensitive to cold than air-source households, consistent with the higher efficiency of ground-source systems at low outdoor temperatures.

## Visual Evidence

The coefficient plot below summarises the direction and significance of all predictors in the extended model. Coefficients whose 95% confidence intervals do not cross zero are statistically distinguishable from no effect.

```{r lm-coef-plot, echo=FALSE, fig.height=7}
cis <- confint(heapo_lm)

coef_df <- tibble(
  term     = names(coef(heapo_lm)),
  estimate = coef(heapo_lm),
  lower    = cis[, 1],
  upper    = cis[, 2]
) %>%
  filter(term != "(Intercept)") %>%
  mutate(significant = !(lower < 0 & upper > 0))

ggplot(coef_df, aes(x = estimate, y = reorder(term, estimate), color = significant)) +
  geom_point(size = 2.5) +
  geom_errorbarh(aes(xmin = lower, xmax = upper), height = 0.3) +
  geom_vline(xintercept = 0, linetype = "dashed", color = "grey50") +
  scale_color_manual(
    values = c("FALSE" = "grey60", "TRUE" = "steelblue"),
    labels = c("n.s.", "p < 0.05")
  ) +
  theme_minimal(base_size = 13) +
  labs(
    title  = "Extended MLR: Coefficient Estimates with 95% CIs",
    subtitle = "Coefficients on log scale — positive = higher consumption",
    x      = "Coefficient estimate",
    y      = NULL,
    color  = NULL
  )
```

The scatterplot of observed versus fitted values on the log scale illustrates the model's predictive fit. Points near the diagonal indicate accurate predictions; systematic deviations would suggest model misspecification.

```{r lm-obs-vs-fit, echo=FALSE, fig.height=5}
fit_df <- tibble(
  fitted   = fitted(heapo_lm),
  observed = heapo_data$log_kWh_total[as.integer(names(fitted(heapo_lm)))]
)

ggplot(fit_df, aes(x = fitted, y = observed)) +
  geom_point(alpha = 0.15, size = 0.8) +
  geom_abline(slope = 1, intercept = 0, color = "steelblue", linetype = "dashed") +
  theme_minimal(base_size = 13) +
  labs(
    title  = "Observed vs. Fitted (log kWh)",
    x      = "Fitted values",
    y      = "Observed values"
  )
```

## Model Diagnostics

```{r lm-diagnostics, echo=FALSE, fig.height=8}
par(mfrow = c(2, 2))
plot(heapo_lm)
par(mfrow = c(1, 1))
```

- **Residuals vs Fitted:** Checks linearity. A horizontal band with no systematic pattern is ideal.
- **Q-Q Plot:** Checks normality of residuals. Points near the diagonal indicate normality; heavy tails would signal departure.
- **Scale-Location:** Checks homoscedasticity (constant variance). A flat, evenly spread line is ideal.
- **Residuals vs Leverage:** Identifies high-leverage observations that disproportionately influence the fit.

```{r lm-vif, echo=FALSE}
vif_vals <- vif(heapo_lm)
data.frame(
  Predictor = names(vif_vals),
  VIF       = round(as.numeric(vif_vals), 2)
) %>%
  knitr::kable(caption = "Variance Inflation Factors (VIF) — extended model")
```

Most predictors have VIF below 3, indicating no problematic multicollinearity. The exception is `heating_degree_days` (VIF = 61.55) and its interaction with `building_typehouse` (VIF = 61.44). These elevated values arise directly from the interaction term — both terms share the same underlying variable — and are therefore expected and do not invalidate the model or inflate coefficient uncertainty in practice.

```{r lm-model-comparison, echo=FALSE}
rmse <- function(model) sqrt(mean(residuals(model)^2))

tibble(
  Model    = c("Simple MLR (main effects)", "Extended MLR (+ interactions)"),
  `Adj. R²` = c(
    round(summary(heapo_mlm_simple)$adj.r.squared, 4),
    round(summary(heapo_lm)$adj.r.squared, 4)
  ),
  RMSE = c(
    round(rmse(heapo_mlm_simple), 4),
    round(rmse(heapo_lm), 4)
  )
) %>%
  knitr::kable(caption = "LM model comparison")
```

The simple model achieves an adjusted R² of 0.2727 (RMSE = 0.8794). Adding the two interaction terms improves this to 0.2743 (RMSE = 0.8785) — a real but modest gain. This confirms the interaction terms are meaningful while also highlighting that the overall explanatory power of a linear model is limited on this dataset.

## Key Findings

**Heating degree days is the dominant weather driver:** Cold weather is the primary positive predictor of energy consumption, with an effect that is further amplified in houses (interaction β = 0.039, p < 0.001) — confirming that houses are thermally less efficient than flats.

**Household size has the strongest marginal effect:** With β = 0.134, each additional resident increases daily consumption by ~14.3%, making it the largest single-predictor effect in the model after the building type interaction.

**Houses consume substantially more, especially in winter:** Both the baseline offset (β = 0.220, house vs. flat) and the temperature interaction (β = 0.039) point in the same direction — houses face higher base consumption and are more sensitive to cold.

**Ground-source heat pumps are more efficient and more stable:** They consume ~13.3% less at baseline and respond less to temperature swings (interaction β = −0.003), consistent with the known efficiency advantage of ground-source systems.

**Humidity is not a useful predictor:** After controlling for other variables, `humidity_avg` is not statistically significant (p = 0.165) and contributes negligible explanatory power.

**Overall explanatory power is limited:** The extended model explains only ~27.4% of the variance in log daily consumption (adj. R² = 0.2743, RMSE = 0.879 on log scale). On the original kWh scale this corresponds to a residual standard error of approximately $e^{0.879} \approx 2.4\times$, meaning predictions can be off by more than a factor of two.

## Conclusion

The linear model successfully identifies the key structural and weather-related drivers of daily household energy consumption and provides clear, interpretable coefficient estimates. All predictors except humidity are statistically significant, and the interaction terms meaningfully capture the heterogeneous effect of cold weather across building and pump types.

However, the model is of **limited usefulness for prediction**. An adjusted R² of 0.27 means the model leaves ~73% of the variance in daily consumption unexplained. At the household-day level, predictions are too imprecise to serve as a reliable forecasting tool. The likely cause is the strong household-level heterogeneity that a cross-sectional linear model cannot capture — different households have very different baseline consumption levels regardless of the measured predictors.

The model is better suited for **inference and understanding drivers** than for prediction. It provides a statistically well-grounded description of which factors are associated with higher consumption and by how much, which is valuable for policy and planning contexts. For operational forecasting, more flexible models with household-level effects (e.g., mixed models, GAMs, or tree-based methods) would be needed.
