---
# title: "Support Vector Machine"
# author: "Tharrmeehan Krishnathasan"
# date: "2026 04 09"
output:
  pdf_document:
    latex_engine: xelatex
    toc: true
    toc_depth: '3'
  html_document:
    toc: true
    toc_float: true
    toc_depth: 3
    number_sections: true
    theme: flatly
    code_folding: hide
subtitle: Applied Machine Learning and Predictive Modelling 1 – FS 2026
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = FALSE
)
library(ggplot2)
library(dplyr)
library(e1071)
library(caret)
library(kernlab)
library(knitr)
set.seed(42)
```
```{r import-svm, echo=FALSE, message=FALSE, warning=FALSE}
svm_caret  <- readRDS("../models/svm/heapo_svm.rds")
heapo_data <- readRDS("../data_processed/heapo/heapo_modelling.rds")

# Reconstruct the classification dataset and test split
q75 <- quantile(heapo_data$log_kWh_total, 0.75)

heapo_svm_data <- heapo_data %>%
  mutate(
    high_consumption = as.factor(
      ifelse(log_kWh_total >= q75, "high", "normal")
    )
  ) %>%
  select(
    high_consumption,
    heating_degree_days,
    sunshine_hours,
    humidity_avg,
    living_area,
    n_residents,
    building_type,
    heatpump_type
  ) %>%
  na.omit()

set.seed(42)
train_idx  <- createDataPartition(heapo_svm_data$high_consumption, p = 0.8, list = FALSE)
train_data <- heapo_svm_data[train_idx, ]
test_data  <- heapo_svm_data[-train_idx, ]
test_x     <- test_data %>% select(-high_consumption)
test_y     <- test_data$high_consumption

# Evaluate the saved tuned model on the test set
pred_best <- predict(svm_caret, test_x)
cm_best   <- confusionMatrix(pred_best, test_y, positive = "high")

# Fit the baseline for comparison (e1071 radial, cost=1, default gamma)
set.seed(42)
svm_baseline  <- svm(
  high_consumption ~ .,
  data   = train_data,
  kernel = "radial",
  cost   = 1,
  scale  = TRUE
)
pred_baseline <- predict(svm_baseline, test_x)
cm_baseline   <- confusionMatrix(pred_baseline, test_y, positive = "high")

# Extract best hyperparameters for inline reporting
best_C     <- svm_caret$bestTune$C
best_sigma <- round(svm_caret$bestTune$sigma, 4)
```

# Support Vector Machine

## Objective

The aim of this analysis is to classify each day as a *high-consumption day* (top 25% of daily kWh usage) or a *normal day* — a binary classification problem. This framing is practically useful: identifying high-load days in advance allows grid operators and energy providers to plan capacity and issue consumption alerts.

A Support Vector Machine (SVM) is employed because it can capture complex, non-linear decision boundaries via kernel functions, without requiring distributional assumptions on the predictors. Model performance is evaluated on a held-out test set using accuracy, sensitivity, specificity, and F1 score.

## Data Preparation

The response variable, `high_consumption`, is derived from `log_kWh_total` using the 75th percentile as a threshold. Days above this threshold are labelled *high*; all others are *normal*. The threshold corresponds to 3.541 on the log scale (approximately 34.5 kWh per day). This data-driven cutoff identifies the 25% of days with the highest energy demand, which are disproportionately responsible for grid stress events.

After removing missing values, the dataset contains 93,686 observations: 23,440 high-consumption days (25.0%) and 70,246 normal days (75.0%). The original `log_kWh_total` is excluded from the predictors to prevent data leakage — the SVM must identify high-load days from weather conditions and household characteristics alone.

The data are split into training (74,949 rows) and test (18,737 rows) sets using stratified 80/20 sampling, preserving the 25/75 class ratio in both subsets. All numeric predictors are standardised before training, as SVMs are sensitive to feature scale.

```{r class-balance, echo=FALSE}
prop.table(table(train_data$high_consumption)) %>%
  as.data.frame() %>%
  rename(Class = Var1, Proportion = Freq) %>%
  mutate(Proportion = round(Proportion, 3)) %>%
  knitr::kable(caption = "Class balance in the training set")
```

## Model Setup

Two SVM models are estimated using a radial basis function (RBF) kernel. The RBF kernel maps data into a high-dimensional space, enabling non-linear decision boundaries — well-suited to energy consumption patterns that interact across weather and household variables.

**Baseline SVM:** Cost = 1, gamma = default (1 / number of features). This serves as a reference point for the tuned model.

**Tuned SVM:** 5-fold cross-validation is used to search over a grid of `C` ∈ {1, 10, 100} and `sigma` ∈ {0.01, 0.1, 1/7}. `C` controls the penalty for misclassification (higher = tighter fit); `sigma` is the inverse bandwidth of the RBF kernel (higher = more local boundaries). Cross-validation is parallelised across cores via `doParallel`. The best parameter combination maximises cross-validation accuracy on the training set.

The tuned model uses the following best parameters: C = 100, sigma = 0.1429 (equal to 1/7, the default gamma in `e1071`). These were selected from a 3×3 grid and achieved the highest cross-validation accuracy of 0.8304 (Kappa = 0.4859).

## Model Results

```{r svm-eval-table, echo=FALSE}
extract_metrics <- function(cm, model_name) {
  tibble(
    Model       = model_name,
    Accuracy    = round(cm$overall["Accuracy"], 4),
    Sensitivity = round(cm$byClass["Sensitivity"], 4),
    Specificity = round(cm$byClass["Specificity"], 4),
    `F1 Score`  = round(cm$byClass["F1"], 4)
  )
}

bind_rows(
  extract_metrics(cm_baseline, "Baseline SVM (C=1, default sigma)"),
  extract_metrics(
    cm_best,
    paste0("Tuned SVM (C=", best_C, ", sigma=", best_sigma, ")")
  )
) %>%
  knitr::kable(caption = "SVM model comparison on test set")
```

The tuned SVM (C = 100, sigma = 0.1429) achieves an accuracy of **83.2%** on the test set, compared to 82.2% for the baseline — a modest improvement. However, accuracy alone is misleading here because a naive classifier that always predicts *normal* would achieve 75.0% (the majority class proportion). The true gain over this baseline is therefore only ~8 percentage points.

The more important metrics for this task are:

**Sensitivity: 0.4834** — the model correctly identifies only 48.3% of actual high-consumption days. More than half of all high-load days are missed. This is a significant limitation for an early-warning application where false negatives carry higher operational costs.

**Specificity: 0.9481** — normal days are classified correctly 94.8% of the time. The model is conservative: it flags fewer high-consumption days than actually occur (1,598 predicted high vs. 2,502 actual high in the test set).

**F1 Score: 0.5899** — the harmonic mean of precision and recall for the high class reflects the imbalance between high specificity and low sensitivity.

**Kappa: 0.4905** — a moderate level of agreement beyond chance, improving on the baseline (0.4602) but still below the 0.6 threshold typically considered *substantial* agreement.

The tuned model is selected as the final model based on the highest cross-validation accuracy on the training set.

## Visual Evidence

The confusion matrix heatmap shows the distribution of predictions versus actual labels. Correct predictions lie on the diagonal; off-diagonal cells represent misclassifications. A concentration of observations on the diagonal confirms strong overall performance.

```{r svm-conf-matrix-plot, echo=FALSE, fig.height=4, fig.width=5}
cm_mat <- as.table(cm_best$table)

cm_df <- as.data.frame(cm_mat) %>%
  rename(Predicted = Prediction, Actual = Reference)

ggplot(cm_df, aes(x = Actual, y = Predicted, fill = Freq)) +
  geom_tile(color = "white") +
  geom_text(aes(label = Freq), size = 5) +
  scale_fill_gradient(low = "#f0f4f8", high = "steelblue") +
  theme_minimal(base_size = 13) +
  labs(
    title = "Confusion Matrix — Tuned SVM",
    fill  = "Count"
  )
```

The bar chart below compares sensitivity and specificity across the baseline and tuned models, illustrating how cross-validation tuning shifts the trade-off between detecting high-consumption days and avoiding false alarms.

```{r svm-metric-comparison, echo=FALSE, fig.height=4, fig.width=7}
metrics_df <- bind_rows(
  extract_metrics(cm_baseline, "Baseline"),
  extract_metrics(cm_best, "Tuned")
) %>%
  select(Model, Sensitivity, Specificity) %>%
  tidyr::pivot_longer(cols = c(Sensitivity, Specificity), names_to = "Metric", values_to = "Value")

ggplot(metrics_df, aes(x = Metric, y = Value, fill = Model)) +
  geom_col(position = "dodge") +
  scale_fill_manual(values = c("Baseline" = "grey70", "Tuned" = "steelblue")) +
  scale_y_continuous(limits = c(0, 1)) +
  theme_minimal(base_size = 13) +
  labs(
    title = "Sensitivity and Specificity: Baseline vs. Tuned SVM",
    x     = NULL,
    y     = "Value",
    fill  = "Model"
  )
```

## Model Diagnostics

SVM diagnostics focus on generalisation performance and the validity of modelling choices rather than distributional assumptions.

**CV tuning results:** Across all 9 combinations, CV accuracy ranged from 0.8071 (C=10, sigma=0.01) to 0.8304 (C=100, sigma=0.1429). The top four combinations all involve sigma ≥ 0.1, suggesting the model benefits from a moderately local kernel. The gain from increasing C from 10 to 100 at sigma=0.1429 is small (0.8268 → 0.8304), indicating diminishing returns from further penalising misclassification.

**Class separability:** Sensitivity of 48.3% reveals that the available predictors (weather conditions and household characteristics) do not reliably separate the two classes. Many high-consumption days fall in the same feature space as normal days, creating an irreducible overlap that no kernel or parameter choice can overcome.

**Tuning stability:** The best parameter combination (C=100, sigma=0.1429) sits at the top-right corner of the grid, raising the question of whether a larger C or sigma would further improve results. However, given the flattening improvement from C=10 to C=100 and the domain knowledge that the classes genuinely overlap, additional tuning is unlikely to produce meaningful sensitivity gains.

**Comparison with baseline:** The tuned model improves accuracy (82.2% → 83.2%) and sensitivity (46.2% → 48.3%) over the default parameterisation. The improvement is real but modest, consistent with the view that the bottleneck is feature informativeness rather than model configuration.

## Key Findings

**The SVM achieves reasonable accuracy but low sensitivity:** At 83.2% accuracy, the model clearly outperforms always predicting the majority class (75.0%). However, sensitivity of 48.3% means it misses more than half of all actual high-consumption days — the most operationally important class.

**High specificity reflects a conservative classifier:** Specificity of 94.8% shows the model rarely raises false alarms on normal days, but this comes at the cost of missing many genuine high-load events.

**Cross-validation tuning yields real but limited improvement:** Moving from default parameters to C=100, sigma=0.1429 improves sensitivity from 46.2% to 48.3% and accuracy from 82.2% to 83.2%. The gain is consistent but small, confirming that further tuning alone cannot resolve the core problem.

**The predictor set is the bottleneck:** Weather conditions and static household characteristics (building type, living area, residents, heat pump type) capture only part of what drives high-consumption days. The strong day-to-day within-household variation that drives a day over the 34.5 kWh threshold is not fully explained by the available features.

**Feature leakage is correctly avoided:** By excluding `log_kWh_total`, all results reflect genuine out-of-sample discriminative ability from operationally observable inputs.

## Conclusion

The SVM demonstrates that weather conditions and household characteristics carry real signal for identifying high-consumption days, achieving ~8 percentage points above the naive majority-class baseline. The model is well-specified and the tuning is sound.

However, the model is of **limited practical usefulness as a standalone early-warning system**. A sensitivity of 48.3% means that in a deployment context, more than half of high-load days would go undetected. For capacity planning or demand-response triggering, this miss rate is likely too high to be operationally reliable.

The root cause is not the model choice — it is the available feature set. High and normal consumption days overlap substantially in the weather-and-household feature space, and no kernel function or regularisation choice can separate what is genuinely overlapping in the data. Meaningful improvement would require richer features: lagged consumption, day-of-week and seasonal indicators, occupancy patterns, or appliance-level data.

If the threshold for flagging is lowered (accepting more false alarms to capture more true positives), the model could play a supporting role in probabilistic forecasting pipelines. But at the default threshold, it should not be used as the sole basis for high-consumption day decisions.
