---
title: "Household Energy Consumption Analysis"
author: "Emanuel Lemma, Raphaël Spagolla, Tharrmeehan Krishnathasan"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: false
    toc_depth: 3
    number_sections: true
    theme: flatly
    code_folding: hide
  pdf_document:
    toc: true
    toc_depth: 3
subtitle: Applied Machine Learning and Predictive Modelling 1 – FS 2026
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = FALSE,
  fig.width = 7,
  fig.height = 4,
  cache = TRUE
)
```

```{css, echo=FALSE}
body {
  font-size: 13px;
  line-height: 1.45;
}
p {
  margin-bottom: 0.55em;
}
h1, h2, h3, h4 {
  margin-top: 1.1em;
  margin-bottom: 0.35em;
}
.main-container {
  padding-top: 10px;
}
```

```{r packages, message=FALSE}
library(dplyr)
library(tidyr)
library(ggplot2)
library(lubridate)
library(knitr)
library(kableExtra)
library(scales)
library(broom) # tidy model output
library(car) # VIF
library(mgcv) # GAM
library(gratia)
library(nnet) # Neural Network
library(ROCR)
library(e1071) # SVM
library(caret) # Cross-validation
library(kernlab) # SVM kernels
library(gridExtra)
```

```{r plot-defaults, include=FALSE}
clr_primary <- "#4E79A7"
clr_secondary <- "#E15759"
clr_neutral <- "#AAAAAA"
clr_builds <- c("appartment" = "#4E79A7", "house" = "#E15759")
report_theme <- theme_minimal(base_size = 12)
```

```{r load-data, include=FALSE}
heapo <- readRDS("../data_processed/heapo/heapo_modelling.rds")
```

# Introduction

With the rapid expansion of heat pump installations across Swiss households,
electricity demand is increasingly driven by heating and cooling needs rather
than fixed consumption habits. Unlike traditional resistive heating, heat pumps
draw substantial power in short, weather-dependent bursts. This creates load
profiles that are harder to predict and more sensitive to outdoor conditions
than the appliance-driven consumption patterns that grid operators have
historically managed.

For energy suppliers and grid operators, this shift has direct economic
consequences. Procurement decisions made a day or week in advance are based on
load forecasts. When a cold snap arrives earlier than expected or a warm spell
pushes temperatures above the heating threshold, actual demand diverges from the
submitted schedule. The resulting imbalances must be settled at spot or
balancing energy prices. These costs ultimately flow back to households and
erode the economic case for electrification if not properly managed.

Against this background, this analysis systematically evaluates smart meter data
from a portfolio of heat pump households in Switzerland, using the **HEAPO
dataset**, in order to identify the most important drivers of electricity
consumption and build a well-founded basis for demand modelling. The focus is
not only on explaining consumption variability across seasonal patterns,
building characteristics, and technology choices, but also on quantifying the
interaction between external factors such as outdoor temperature, heating degree
days, and building type.

The report is structured as follows. Section 2 presents exploratory data
analysis. Section 3 fits and interprets six models, namely a Linear Model,
GLM Poisson, GLM Binomial, Generalised Additive Model, Neural Network, and
Support Vector Machine. Section 4 compares model performance. Section 5
draws conclusions for energy suppliers and grid operators.


## Dataset

**Source:** HEAPO – Household Energy and Smart Meter Data (Zenodo, 2025).

The HEAPO dataset was collected as part of a Swiss research project monitoring
residential heat pump installations via smart meters. It covers `r nrow(heapo)`
household-days across `r length(unique(heapo$Household_ID))` households observed
between 2022 and 2024, paired with daily weather data from the corresponding
measurement region. Each observation represents one household on one calendar
day, with daily electricity consumption as the primary outcome alongside a
set of building, technology, and weather covariates.

## Variable Description

| Variable | Type | Description |
|:---|:---|:---|
| `log_kWh_total` | Continuous | Natural log of daily total electricity consumption (kWh). Response for LM and GAM. |
| `heating_degree_days` | Continuous | Daily heating demand below the 12°C Swiss SIA base temperature (°C·day). |
| `temp_avg` | Continuous | Daily average outdoor temperature (°C). |
| `living_area` | Continuous | Heated floor area of the household (m²). |
| `building_type` | Categorical (>2 levels) | Building type, either "appartment" or "house". |
| `heatpump_type` | Categorical (2 levels) | Heat pump technology: air-source or ground-source. |
| `n_residents` | Count | Number of permanent household members. |
| `is_weekend` | Binary | 1 if Saturday or Sunday, 0 otherwise. |
| `has_floor_heating` | Binary | 1 if floor heating system present, 0 otherwise. |
| `high_consumption` | Binary | 1 if daily kWh exceeds the household's own 75th percentile (top 25% of that household's days). Response for GLM Binomial. |

The raw HEAPO dataset spans 2018 to 2024 across 438 monitored households in the
canton of Zurich. The modelling subset was restricted to the period 2022-01-01
through 2024-02-29 and limited to households with at least 95% daily coverage to
exclude partially observed installations. It was then randomly sampled to a
maximum of 100,000 observations to comply with the course data size limit.

# Exploratory Data Analysis

Daily electricity consumption (`kWh_total`) is right-skewed on the raw scale and
approximately symmetric after log transformation. The log-transformed variable
`log_kWh_total` is used as the continuous response in all regression models. The
binary variable `high_consumption` is derived from it. A day is classified as
high consumption if daily `log_kWh_total` exceeds that household's own 75th
percentile, flagging the top 25% of consumption days for each individual
household.

## Seasonal Pattern

```{r eda-seasonal, fig.height=4, fig.align='center'}
heapo %>%
  mutate(Month = month(date, label = TRUE, abbr = TRUE)) %>%
  ggplot(aes(x = Month, y = kWh_total)) +
  geom_boxplot(fill = clr_primary, alpha = 0.7, outlier.size = 0.2) +
  scale_y_continuous(labels = comma) +
  labs(
    title = "Daily Electricity Consumption by Month",
    x = NULL,
    y = "kWh per day"
  ) +
  report_theme
```

Consumption in winter months (December to February) is roughly three to five
times higher than in summer (June to August), driven almost entirely by heat
pump operation for space heating. The wide interquartile range in winter
reflects differences in building size, insulation quality, and heat pump
technology across households. This strong seasonal pattern motivates heating
degree days as the primary weather predictor in all models.

## Heating Degree Days vs. Consumption

**Heating Degree Days (HDD)** quantify daily heating demand as the temperature
shortfall below the 12°C Swiss SIA base temperature. A day averaging 2°C
contributes 10 HDD, days above 12°C contribute zero. HDD is a more direct proxy
for heat pump workload than raw outdoor temperature.

```{r eda-hdd, fig.height=4, fig.align='center'}
heapo %>%
  slice_sample(n = 15000) %>%
  ggplot(aes(x = heating_degree_days, y = kWh_total)) +
  geom_point(alpha = 0.08, size = 0.5, color = clr_primary) +
  geom_smooth(
    method = "loess",
    color = clr_secondary,
    se = FALSE,
    linewidth = 1
  ) +
  scale_y_continuous(labels = comma) +
  labs(
    title = "Heating Degree Days vs. Daily Consumption",
    subtitle = "Random subsample of 15,000 observations",
    x = "Heating Degree Days (HDD)",
    y = "kWh per day"
  ) +
  report_theme
```

The relationship is positive and seams to be linear across the observed
HDD range, with consumption measured form 0 to 125 kWh per day at 0 heating degree days
, which hints to the base load of the building types. The loess smoother confirms that no sharp threshold or strong
curvature is present, which justifies treating HDD as a continuous predictor in
the Linear Model. A slight levelling off at very high HDD values is visible, and
the GAM will capture this through its flexible smooth terms.

## Consumption by Building Type

```{r eda-building, fig.height=3.5, fig.align='center'}
heapo %>%
  ggplot(aes(x = building_type, y = kWh_total, fill = building_type)) +
  geom_boxplot(alpha = 0.7, outlier.size = 0.2) +
  scale_y_continuous(labels = comma) +
  scale_fill_manual(values = clr_builds, guide = "none") +
  labs(
    title = "Daily Consumption by Building Type",
    x = NULL,
    y = "kWh per day"
  ) +
  report_theme
```

Houses consume substantially more electricity per day than apartments, with medians
differing by roughly a factor of two. This gap persists after controlling for
weather, as the modelling section will confirm. Building type therefore carries
real structural signal and is included as a predictor in all models.

Taken together, the three plots tell a consistent story. Electricity demand is
primarily a seasonal phenomenon. The portfolio shifts from a low-consumption
summer baseline driven by domestic appliances to a high-consumption winter
regime dominated by heat pump space heating. Within that seasonal pattern,
building type is the strongest structural separator, with houses drawing
roughly twice the daily energy of apartments at any given temperature. These
findings directly shape the modelling choices below. The response is
log-transformed to stabilise variance, heating degree days enter as the
primary continuous weather predictor, and building type is retained as a
categorical covariate in every model. For the client, the key implication is
that cold spells and the building stock composition of a supply portfolio are
the two dominant levers of aggregate demand, and any forecasting tool must
capture both to be useful.

# Modelling

<!-- Each model section must:
       1. Fit the model to an appropriate response variable
       2. Interpret the coefficients in plain language (not just report p-values)
       3. State which group member led that section (at least 3 of the 6 models need a named lead)
     Models are saved as .RDS files and reloaded to speed up compilation. -->

Six models are fitted to the HEAPO data, each targeting a response variable
suited to its statistical family. The Linear Model and Generalised Additive
Model are fitted to `log_kWh_total`, the log-transformed daily electricity
consumption. The GLM Binomial, Neural Network, and Support Vector Machine all classify
household-days as high or low consumption using each household's own 75th
percentile as the threshold, capturing the top 25% of days for each individual
household. The GLM Poisson
models the monthly count of high-consumption days per household, capturing
longer-term demand patterns at a coarser aggregation. Model performance is
evaluated in the comparison section: regression models are compared via
five-fold cross-validation MAE in kWh and R²; classification models via AUC
and accuracy.

## Linear Model (LM) {#lm}

*Lead: Tharrmeehan Krishnathasan*

```{r import-lm, echo=TRUE, message=FALSE, warning=FALSE}
heapo_lm <- readRDS("../models/linear_model/heapo_lm.rds")

mlm_simple_path <- "../models/linear_model/heapo_mlm_simple.rds"
if (file.exists(mlm_simple_path)) {
  heapo_mlm_simple <- readRDS(mlm_simple_path)
} else {
  heapo_mlm_simple <- lm(
    log_kWh_total ~ heating_degree_days +
      sunshine_hours +
      humidity_avg +
      living_area +
      n_residents +
      building_type +
      heatpump_type,
    data = heapo
  )
  saveRDS(heapo_mlm_simple, mlm_simple_path)
}
```

### Objective

This section models daily household energy consumption as $\text{log\_kWh\_total}$, the natural logarithm of total daily kilowatt-hour usage. The log transformation is standard for energy data: daily kWh is strictly positive and right-skewed in raw form, and approximately normal on the log scale.

Predictors cover weather conditions (heating degree days, sunshine hours, humidity), building characteristics (building type, living area, heat pump type), and household composition (number of residents). Two interaction terms are added to test whether cold-weather sensitivity differs by building type and heat pump technology.

### Model Setup

Two models are estimated:

**Simple MLR:** all seven predictors with main effects only, as an interpretable baseline for the additive relationships between predictors and log energy consumption:

$$
\log(\text{kWh}_i) = \beta_0 + \beta_1\,\text{HDD}_i + \beta_2\,\text{Sunshine}_i + \cdots + \beta_7\,\text{HeatpumpType}_i + \varepsilon_i
$$

**Extended MLR:** adds two interaction terms motivated by domain knowledge:

- $\text{heating\_degree\_days} \times \text{heatpump\_type}$: different heat
  pump technologies have different efficiencies in cold conditions, so the
  effect of cold on consumption should vary by pump type.
- $\text{heating\_degree\_days} \times \text{building\_type}$: houses and
  apartments differ structurally in size, exposed surface area, and construction,
  so their cold-weather sensitivity is expected to differ.

All coefficients are on the log scale: a coefficient of 0.05 means a 5%
multiplicative change in kWh for a one-unit increase in the predictor.

The panel structure (multiple observations per household) means the i.i.d. assumption of standard OLS is not fully met; standard errors may be slightly understated due to within-household correlation.

### Model Results

<details>
<summary>Show / hide table</summary>
```{r lm-coef-table, echo=TRUE}
tidy(heapo_lm) %>%
  filter(term != "(Intercept)") %>%
  mutate(
    estimate = round(estimate, 4),
    std.error = round(std.error, 4),
    statistic = round(statistic, 2),
    p.value = signif(p.value, 3)
  ) %>%
  knitr::kable(
    col.names = c("Predictor", "Estimate", "Std. Error", "t value", "p value"),
    caption = "Extended MLR: coefficient estimates"
  ) %>%
  kable_styling(bootstrap_options = c("striped", "hover"), full_width = TRUE)
```
</details>

All predictors are significant at the 0.001 level except `humidity_avg` (p = 0.165), which adds nothing once the other weather variables are included. Effects on log energy consumption, holding all other variables constant:

**Heating degree days** (β = 0.021, p < 0.001): Each additional HDD increases log consumption by ~2.1% for air-source pump households in apartments (the baseline group).

**Sunshine hours** (β = −0.015, p < 0.001): An additional sunshine hour reduces
consumption by ~1.5%, reflecting lower heating demand on brighter days.

**Number of residents** (β = 0.134, p < 0.001): Each additional resident
increases consumption by ~14.3% ($e^{0.134} - 1$), the second strongest effect
after the temperature-building interaction.

**Building type — house** (β = 0.220, p < 0.001): Houses consume ~24.6% more than apartments at baseline, before accounting for their greater sensitivity to cold (see interaction below).

**Heat pump type — ground-source** (β = −0.143, p < 0.001): Ground-source pump
households consume ~13.3% less than air-source households at the baseline
temperature.

**Interaction: HDD × house** (β = 0.039, p < 0.001): Houses are substantially *more* sensitive to cold than apartments: each additional HDD adds 3.9% to log consumption, the largest interaction effect in the model.

**Interaction: HDD × ground-source** (β = −0.003, p < 0.001): Ground-source pump
households are slightly *less* sensitive to cold than air-source households,
consistent with the higher efficiency of ground-source systems at low outdoor
temperatures.

### Visual Evidence

The coefficient plot below summarises the direction and significance of all predictors in the extended model.

```{r lm-coef-plot, echo=TRUE, fig.height=5, fig.width=8, out.width="100%", fig.align="center"}
cis <- confint(heapo_lm)

coef_df <- tibble(
  term = names(coef(heapo_lm)),
  estimate = coef(heapo_lm),
  lower = cis[, 1],
  upper = cis[, 2]
) %>%
  filter(term != "(Intercept)") %>%
  mutate(significant = !(lower < 0 & upper > 0))

ggplot(
  coef_df,
  aes(x = estimate, y = reorder(term, estimate), color = significant)
) +
  geom_point(size = 2.5) +
  geom_errorbarh(aes(xmin = lower, xmax = upper), height = 0.3) +
  geom_vline(xintercept = 0, linetype = "dashed", color = clr_neutral) +
  scale_color_manual(
    values = c("FALSE" = clr_neutral, "TRUE" = clr_primary),
    labels = c("n.s.", "p < 0.05")
  ) +
  report_theme +
  labs(
    title = "Linear Model – Coefficient Estimates with 95% CIs",
    subtitle = "Coefficients on log scale — positive = higher consumption",
    x = "Coefficient estimate",
    y = NULL,
    color = NULL
  )
```

The scatterplot of observed versus fitted values on the log scale illustrates the model's predictive fit. Points near the diagonal indicate accurate predictions.

```{r lm-obs-vs-fit, echo=TRUE, fig.height=3.5, fig.width=5.5, out.width="70%", fig.align="center"}
fit_df <- tibble(
  fitted = fitted(heapo_lm),
  observed = heapo$log_kWh_total[as.integer(names(fitted(heapo_lm)))]
)

ggplot(fit_df, aes(x = fitted, y = observed)) +
  geom_point(alpha = 0.15, size = 0.8, color = clr_primary) +
  geom_abline(
    slope = 1,
    intercept = 0,
    color = clr_secondary,
    linetype = "dashed"
  ) +
  report_theme +
  labs(
    title = "Observed vs. Fitted (log kWh)",
    x = "Fitted values",
    y = "Observed values"
  )
```

### Model Diagnostics

```{r lm-diagnostics, echo=TRUE, fig.height=5, fig.width=7, fig.align='center'}
aug <- broom::augment(heapo_lm)

p1 <- ggplot(aug, aes(x = .fitted, y = .resid)) +
  geom_point(alpha = 0.1, size = 0.5, color = clr_primary) +
  geom_smooth(
    method = "loess",
    se = FALSE,
    color = clr_secondary,
    linewidth = 0.8
  ) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "grey50") +
  labs(title = "Residuals vs Fitted", x = "Fitted values", y = "Residuals") +
  report_theme

p2 <- ggplot(aug, aes(sample = .std.resid)) +
  stat_qq(alpha = 0.2, size = 0.5, color = clr_primary) +
  stat_qq_line(color = clr_secondary, linetype = "dashed") +
  labs(
    title = "Normal Q-Q",
    x = "Theoretical quantiles",
    y = "Standardised residuals"
  ) +
  report_theme

p3 <- ggplot(aug, aes(x = .fitted, y = sqrt(abs(.std.resid)))) +
  geom_point(alpha = 0.1, size = 0.5, color = clr_primary) +
  geom_smooth(
    method = "loess",
    se = FALSE,
    color = clr_secondary,
    linewidth = 0.8
  ) +
  labs(
    title = "Scale-Location",
    x = "Fitted values",
    y = expression(sqrt("|Std. residuals|"))
  ) +
  report_theme

p4 <- ggplot(aug, aes(x = .hat, y = .std.resid)) +
  geom_point(alpha = 0.2, size = 0.5, color = clr_primary) +
  geom_smooth(
    method = "loess",
    se = FALSE,
    color = clr_secondary,
    linewidth = 0.8
  ) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "grey50") +
  labs(
    title = "Residuals vs Leverage",
    x = "Leverage",
    y = "Standardised residuals"
  ) +
  report_theme

gridExtra::grid.arrange(p1, p2, p3, p4, ncol = 2)
```

<details>
<summary>Show / hide table</summary>
```{r lm-vif, echo=TRUE}
vif_vals <- vif(heapo_lm)
data.frame(
  Predictor = names(vif_vals),
  VIF = round(as.numeric(vif_vals), 2)
) %>%
  knitr::kable(
    caption = "Variance Inflation Factors (VIF) — extended model"
  ) %>%
  kable_styling(bootstrap_options = c("striped", "hover"), full_width = TRUE)
```
</details>

Most predictors have VIF below 3, indicating no problematic multicollinearity.
The exception is `heating_degree_days` (VIF = 61.55) and its interaction with
`building_typehouse` (VIF = 61.44). These elevated values are expected for interaction terms and do not invalidate the model or inflate coefficient uncertainty.

```{r lm-cv, echo=TRUE, cache=TRUE, message=FALSE, warning=FALSE}
df_lm_cv <- heapo %>%
  select(
    log_kWh_total,
    heating_degree_days,
    sunshine_hours,
    humidity_avg,
    living_area,
    n_residents,
    building_type,
    heatpump_type
  ) %>%
  na.omit() %>%
  as.data.frame()

set.seed(42)
folds_lm <- caret::createFolds(df_lm_cv$log_kWh_total, k = 5, list = TRUE)

cv_lm <- lapply(folds_lm, function(test_idx) {
  train <- df_lm_cv[-test_idx, ]
  test <- df_lm_cv[test_idx, ]

  simple_fit <- lm(
    log_kWh_total ~ heating_degree_days +
      sunshine_hours +
      humidity_avg +
      living_area +
      n_residents +
      building_type +
      heatpump_type,
    data = train
  )
  ext_fit <- lm(
    log_kWh_total ~ heating_degree_days *
      building_type +
      heating_degree_days * heatpump_type +
      sunshine_hours +
      humidity_avg +
      living_area +
      n_residents,
    data = train
  )

  actual_kwh <- exp(test$log_kWh_total)
  list(
    simple_mae = mean(abs(
      actual_kwh - exp(predict(simple_fit, newdata = test))
    )),
    ext_mae = mean(abs(actual_kwh - exp(predict(ext_fit, newdata = test))))
  )
})

lm_simple_cv_mae <- mean(sapply(cv_lm, `[[`, "simple_mae"))
lm_ext_cv_mae <- mean(sapply(cv_lm, `[[`, "ext_mae"))
```

<details>
<summary>Show / hide table</summary>
```{r lm-model-comparison, echo=TRUE}
tibble(
  Model = c("Simple MLR (main effects)", "Extended MLR (+ interactions)"),
  `Adj. R²` = c(
    round(summary(heapo_mlm_simple)$adj.r.squared, 4),
    round(summary(heapo_lm)$adj.r.squared, 4)
  ),
  `CV MAE (kWh/day)` = c(
    round(lm_simple_cv_mae, 2),
    round(lm_ext_cv_mae, 2)
  )
) %>%
  knitr::kable(
    caption = "LM model comparison — Adj. R² on full data; MAE from 5-fold CV"
  ) %>%
  kable_styling(bootstrap_options = c("striped", "hover"), full_width = TRUE)
```
</details>

The simple model achieves an adjusted R² of `r round(summary(heapo_mlm_simple)$adj.r.squared, 4)`. Adding the two interaction terms improves this to `r round(summary(heapo_lm)$adj.r.squared, 4)`, a real but modest gain. The interaction terms are meaningful, but overall explanatory power remains limited on this dataset.

### Key Findings

**Heating degree days is the dominant weather driver:** Cold weather is the primary positive predictor of energy consumption. Houses start from a higher baseline (β = 0.220) and are more sensitive to cold than apartments (interaction β = 0.039, p < 0.001), consistent with lower thermal efficiency.

**Household size has the strongest marginal effect:** With β = 0.134, each additional resident increases daily consumption by ~14.3%, making it the largest single-predictor effect in the model after the building type interaction.

**Ground-source heat pumps are more efficient and more stable:** They consume ~13.3% less at baseline and respond less to temperature swings (interaction β = −0.003), consistent with the efficiency advantage of ground-source systems at low outdoor temperatures.

**Humidity is not a useful predictor:** `humidity_avg` is not significant (p = 0.165) once the other weather variables are controlled for.

**Overall explanatory power is limited:** The extended model explains ~`r round(summary(heapo_lm)$adj.r.squared * 100, 1)`% of the variance in log daily consumption (adj. R² = `r round(summary(heapo_lm)$adj.r.squared, 4)`). On the original kWh scale, mean absolute error is `r round(lm_ext_cv_mae, 1)` kWh/day (5-fold CV). Household-level heterogeneity is the likely cause; models with household-level random effects would be needed for operational forecasting.

## GLM – Poisson {#glm-poisson}

*Lead: Raphaël Spagolla*

```{r import-poisson, echo=FALSE, cache=FALSE}
dat_count <- readRDS("../data_processed/poisson/dat_count.rds")
mod_pois <- readRDS("../models/poisson_glm/mod_pois.rds")
mod_quasi_pois <- readRDS("../models/poisson_glm/mod_quasi.rds")
```

### Objective

The aim of this analysis is to model the number of high-consumption days per
household-month ($n_{high\ days}$). This variable represents count data, taking
non-negative integer values. A plain linear model is not appropriate here, as it
assumes normally distributed errors and can produce negative predictions — both
of which are incompatible with count outcomes.

A Poisson Generalised Linear Model (GLM) is used instead, as it is specifically
designed for non-negative integer outcomes and naturally allows the variance to
grow with the mean. By including an offset for the number of observed days, the
model estimates rates, allowing interpretation of results in terms of relative
changes (rate ratios).

### Model Setup
<details>
<summary>Show / hide result</summary>
```{r poisson-modell, echo=TRUE}
print(formula(mod_pois))
```
</details>


The response variable is the number of high-consumption days per household-month
($n_{high\ days}$). A day is classified as a high-consumption day if its daily
`log_kWh_total` exceeds that household's own 75th percentile, consistent with
the threshold used in the GLM Binomial and throughout the report.
Because the number of observed days can vary slightly across
household–months, the model includes the logarithm of the observed days
$(\log(n_{days}))$ as an offset. This adjustment effectively converts the
outcome into a rate — high-consumption days per observed day — so that
observations remain comparable even when exposure differs between months.

To model these rates, a Poisson GLM with a log-link is fitted:

$$
\log(\mathbb{E}[Y_i]) = \beta_0 + \beta_1 X_{1i} + \cdots + \beta_p X_{pi} + \log(n\_days_i)
$$

The log-link guarantees positive fitted values and supports a multiplicative
interpretation of effects.

One predictor warrants clarification: `share_weekend` is the monthly proportion
of weekend days within a household–month (a continuous value between 0 and 1),
and is distinct from the daily binary `is_weekend` flag used in the GLM Binomial
and other daily-level models.

### Model Results

The results are interpreted in terms of *percentage changes in the expected
rate*, holding all other variables constant.

**Building type (house vs apartments):** Households living in houses have a rate of
high-consumption days approximately 363% higher than those living in apartments
$((4.632 - 1) \times 100\%)$. This is by far the strongest effect in the model,
clearly visible in the boxplot, and indicates that building type is the dominant
structural driver of consumption patterns.

**Average temperature:** The rate ratio of 0.915 implies that each additional
degree Celsius is associated with an 8.5% decrease in the expected rate of high-
consumption days $((1 - 0.915) \times 100\%)$. This reflects reduced heating
needs at warmer temperatures, consistent with the strong negative trend visible
in the scatterplot.

**Heating degree days:** The rate ratio is marginally below 1 ($\approx$ 0.999)
and not statistically significant. Once average temperature is already in the
model, heating degree days add little additional explanatory power, as both
variables capture overlapping thermal information. This collinearity explains
why the coefficient is effectively zero.

**Number of residents:** Each additional household member increases the expected
rate by approximately 8.9% $((1.089 - 1) \times 100\%)$. This reflects the
intuitive result that larger households consume more energy overall.

**Living area:** The rate ratio of 1.001 indicates a statistically significant
but practically small positive effect — each additional square metre increases
the expected rate by about 0.1%. The effect accumulates for larger dwellings but
remains modest compared to building type or household size.

**Ground-source heat pump:** Households equipped with a ground-source heat pump
show a rate approximately 9.3% lower than the reference heat pump type $((1 -
0.907) \times 100\%)$. This is consistent with ground-source systems being more
energy-efficient, particularly during colder periods.

**Floor heating:** Households with floor heating show a rate approximately 13%
higher than those without $((1.130 - 1) \times 100\%)$. This likely reflects the
greater energy input required by radiant floor systems to maintain comfort
temperatures.

**Weekend share:** This predictor has a very wide confidence interval and is
non-significant once overdispersion is accounted for (quasi-Poisson p > 0.05).
The direction of the raw Poisson estimate should not be interpreted. The two
models are not directly comparable on this predictor, as they operate at
different aggregation levels.

Overall, structural and household characteristics (building type, size, and
number of residents) exhibit the strongest and most reliable effects, while
temperature captures an important environmental influence.

### Visual Evidence

To complement the model output, the plots below provide direct visual support
for the main relationships in the data.

The *boxplot by building type* shows a striking separation between apartments
and houses. Apartments cluster tightly with a median close to 3 high-consumption
days and only two mild outliers, while houses display a median around 16 days
with a very wide interquartile range extending from nearly 0 to 28. This large
difference in both level and spread confirms that building type is the dominant
predictor in the model.

The *scatterplot of high-consumption days versus average temperature* reveals a
strong and consistent negative pattern. At temperatures near 0°C, many
households record close to 30 high-consumption days, whereas above 15°C the
counts concentrate near zero. The smooth LOESS curve decreases steeply in the
0–10°C range and flattens at higher temperatures, suggesting diminishing
marginal effects of warmth once heating demand is already low.

```{r glm-poisson-plots, echo=TRUE, fig.height=4, fig.show='hold', out.width='50%'}
ggplot(dat_count, aes(x = building_type, y = n_high_days, fill = building_type)) +
  geom_boxplot(alpha = 0.7, outlier.size = 0.2) +
  scale_fill_manual(values = clr_builds, guide = "none") +
  labs(
    title = "High-consumption days by building type",
    x = "Building type",
    y = "Number of high-consumption days"
  ) +
  report_theme

ggplot(dat_count, aes(x = temp_avg, y = n_high_days)) +
  geom_point(alpha = 0.3) +
  geom_smooth(method = "loess", color = clr_secondary) +
  labs(
    title = "High-consumption days vs temperature",
    x = "Average temperature",
    y = "High-consumption days"
  ) +
  report_theme
```

The *rate-ratio plot* below summarises all model effects on a log scale. *House
(vs Apartments)* stands far to the right of the reference line (RR $\approx$ 4.6),
confirming it as the dominant effect. *Weekend share* lies furthest to the left
but with a very wide CI crossing 1, indicating high uncertainty. *Temperature*
and *ground-source heat pump* show clear reductions (RR < 1), while *floor
heating*, *number of residents*, and *living area* show moderate to small
increases. *Heating degree days* sits right at 1 and is the only non-significant
predictor (shown in grey).

```{r glm-poisson-rr-plot, echo=TRUE, fig.height=5, fig.width=7, fig.align='center'}
coef_summary <- coef(summary(mod_pois))

coef_df <- data.frame(
  term = rownames(coef_summary),
  estimate = coef_summary[, "Estimate"],
  se = coef_summary[, "Std. Error"],
  p_value = coef_summary[, "Pr(>|z|)"]
)

coef_df <- coef_df[coef_df$term != "(Intercept)", ]

coef_df$rate_ratio <- exp(coef_df$estimate)
coef_df$ci_low <- exp(coef_df$estimate - 1.96 * coef_df$se)
coef_df$ci_high <- exp(coef_df$estimate + 1.96 * coef_df$se)
coef_df$significant <- coef_df$p_value < 0.05

coef_df$term <- dplyr::recode(
  coef_df$term,
  "temp_avg" = "Temperature",
  "living_area" = "Living area",
  "building_typehouse" = "House (vs Apartments)",
  "heatpump_typeground-source" = "Ground-source heat pump",
  "has_floor_heating1" = "Floor heating",
  "n_residents" = "Number of residents",
  "share_weekend" = "Weekend share",
  "heating_degree_days" = "Heating degree days"
)

ggplot(
  coef_df,
  aes(x = rate_ratio, y = reorder(term, rate_ratio), color = significant)
) +
  geom_point(size = 2.5) +
  geom_errorbarh(aes(xmin = ci_low, xmax = ci_high), height = 0.3) +
  geom_vline(xintercept = 1, linetype = "dashed", color = clr_neutral) +
  scale_color_manual(
    values = c("FALSE" = clr_neutral, "TRUE" = clr_primary),
    labels = c("n.s.", "p < 0.05")
  ) +
  scale_x_log10() +
  labs(
    title = "GLM Poisson – Rate Ratios with 95% CIs",
    x = "Rate Ratio (log scale)",
    y = NULL,
    color = NULL
  ) +
  report_theme
```

### Model Diagnostics

```{r diagnostics, echo=TRUE}
s_m_p <- summary(mod_pois)
disp_pois <- s_m_p$deviance / s_m_p$df.residual

s_m_q <- summary(mod_quasi_pois)
phi_q <- sum(residuals(mod_quasi_pois, type = "pearson")^2) /
  df.residual(mod_quasi_pois)

p_pois <- coef(summary(mod_pois))[, 4]
p_quasi <- coef(summary(mod_quasi_pois))[, 4]
```

A key assumption of the Poisson GLM is *equidispersion*, i.e., that the
conditional variance equals the conditional mean. We assess this by comparing
the residual deviance to the residual degrees of freedom. For the fitted Poisson
model, the residual deviance is `r format(round(s_m_p$deviance, 1), big.mark = ",", scientific = FALSE)` on `r s_m_p$df.residual` df, yielding a ratio of `r round(disp_pois, 2)`. Since this
substantially exceeds 1, there is strong evidence of *overdispersion* — the
Poisson model underestimates standard errors and overstates statistical
significance.

We therefore refit the model using a *quasi-Poisson* family, which retains
identical point estimates but inflates standard errors by $\sqrt{\hat{\phi}}$
($\hat{\phi}$ = `r round(phi_q, 2)`), yielding more conservative and reliable
inference. The most visible consequence is that *weekend share* shifts from
significant under Poisson (p = `r signif(p_pois["share_weekend"], 2)`) to
clearly non-significant under quasi-Poisson (p = `r signif(p_quasi["share_weekend"], 2)`), indicating that this effect should not be
over-interpreted.

The table and plot below show the SE inflation across all predictors. The effect
is proportional — predictors with larger SEs to begin with (e.g., *weekend
share*, *house vs apartments*) show the largest absolute increase, while those with
very small SEs (e.g., *living area*, *number of residents*) are inflated by a
smaller absolute amount but the same factor $\sqrt{\hat{\phi}}$.

```{r quasi-se-plot, echo=TRUE, fig.height=4, fig.align='center'}
pois_tidy <- tidy(mod_pois) %>% select(term, estimate, std.error, p.value)
quasi_tidy <- tidy(mod_quasi_pois) %>% select(term, std.error, p.value)

comp_table <- pois_tidy %>%
  filter(term != "(Intercept)") %>%
  rename(se_pois = std.error, pval_pois = p.value) %>%
  left_join(
    quasi_tidy %>%
      filter(term != "(Intercept)") %>%
      rename(se_quasi = std.error, pval_quasi = p.value),
    by = "term"
  ) %>%
  mutate(
    term = dplyr::recode(
      term,
      "temp_avg" = "Temperature",
      "living_area" = "Living area",
      "building_typehouse" = "House (vs apartments)",
      "heatpump_typeground-source" = "Ground-source heat pump",
      "has_floor_heating1" = "Floor heating",
      "n_residents" = "Number of residents",
      "share_weekend" = "Weekend share",
      "heating_degree_days" = "Heating degree days"
    ),
    se_pois = round(se_pois, 4),
    se_quasi = round(se_quasi, 4)
  )

comp_long <- comp_table %>%
  mutate(se_diff = se_quasi - se_pois) %>%
  arrange(se_diff) %>%
  mutate(term = factor(term, levels = term)) %>%
  pivot_longer(
    cols = c(se_pois, se_quasi),
    names_to = "model",
    values_to = "se"
  ) %>%
  mutate(model = ifelse(model == "se_pois", "Poisson", "Quasi-Poisson"))

ggplot(comp_long, aes(x = se, y = term, colour = model)) +
  geom_line(aes(group = term), colour = clr_neutral, linewidth = 0.5) +
  geom_point(size = 2.5) +
  scale_colour_manual(
    values = c("Poisson" = clr_primary, "Quasi-Poisson" = clr_secondary)
  ) +
  geom_vline(xintercept = 0, linetype = "dashed", colour = "grey50") +
  labs(
    title = "Standard errors: Poisson vs quasi-Poisson",
    x = "Standard error",
    y = NULL,
    colour = "Model"
  ) +
  report_theme
```

### Goodness of Fit

To assess predictive accuracy, we compare fitted values against observed counts
and report the Root Mean Squared Error (RMSE). The observed-vs-fitted plot is
rendered as a hexbin to handle overplotting given the large number of
observations.

<details>
<summary>Show / hide result</summary>
```{r goodness-of-fit, echo=TRUE, fig.height=4}
dat_count$fitted_pois <- fitted(mod_pois)
dat_count$fitted_quasi <- fitted(mod_quasi_pois)

rmse_pois <- sqrt(mean((dat_count$n_high_days - dat_count$fitted_pois)^2))
rmse_quasi <- sqrt(mean((dat_count$n_high_days - dat_count$fitted_quasi)^2))

knitr::kable(
  data.frame(
    Model = "Poisson",
    RMSE = round(rmse_pois, 3)
  ),
  caption = "Root Mean Squared Error for the Poisson model."
) %>%
  kable_styling(bootstrap_options = c("striped", "hover"), full_width = TRUE)
```
</details>

The RMSE of `r round(rmse_pois, 3)` days reflects substantial residual spread, partly explained by the high overdispersion ($\hat{\phi}$ = `r round(phi_q, 2)`). Note that the quasi-Poisson model produces identical fitted values and therefore the same RMSE, as it only affects standard errors and inference, not point predictions.

### Key Findings

The Poisson GLM provides a clear and interpretable account of the factors
driving the frequency of high-consumption days. Building type, temperature, and
household size emerge as the strongest and most robust predictors, consistent
with both domain knowledge and the exploratory analysis.

However, a dispersion ratio of `r round(disp_pois, 2)` confirms substantial
overdispersion, making standard Poisson inference unreliable. The quasi-Poisson
model corrects for this by inflating standard errors appropriately, leading to
more conservative and trustworthy conclusions — most notably, the loss of
significance for weekend share.

While the quasi-Poisson model improves inference, a negative binomial model
could be considered as a more flexible alternative, as it models the
overdispersion through an explicit distributional parameter rather than a post-
hoc correction.

## GLM – Binomial {#glm-binomial}

*Lead: Emanuel Lemma*

### Objective

Can we predict whether a given day will be unusually high-consumption for a
specific household? This section fits a **Generalised Linear Model with a
binomial family** to answer exactly that question. The response variable
`high_consumption` is a binary indicator that equals 1 if the household's daily
`log_kWh_total` exceeds that household's own 75th percentile and 0 otherwise.
Using a per-household threshold means the model is not distinguishing houses
from apartments. Each household's own consumption baseline is already accounted for.
Instead, the model learns what drives a day to be in the top 25% for a given
household, regardless of its size or type. By modelling the log-odds of high
consumption as a linear function of weather conditions, building characteristics,
and calendar effects, we obtain a model that is both interpretable and actionable
for a client interested in understanding what drives demand spikes at the
household level.

### Model Setup

```{r glm-binomial-prep}
set.seed(42)

hh_q75 <- heapo |>
  group_by(Household_ID) |>
  summarise(
    hh_q75 = quantile(log_kWh_total, 0.75, na.rm = TRUE),
    .groups = "drop"
  )

df_glm <- heapo |>
  left_join(hh_q75, by = "Household_ID") |>
  mutate(
    high_consumption = as.integer(log_kWh_total >= hh_q75),
    month = as.factor(month)
  ) |>
  select(
    high_consumption,
    heating_degree_days,
    temp_avg,
    living_area,
    building_type,
    heatpump_type,
    has_floor_heating,
    n_residents,
    is_weekend,
    month
  ) |>
  drop_na()

train_idx <- sample(nrow(df_glm), 0.8 * nrow(df_glm))
train_df <- df_glm[train_idx, ]
test_df <- df_glm[-train_idx, ]
```

```{r glm-binomial-fit, cache=TRUE}
if (!file.exists("../models/glm_binomial/glm_binomial_hh_q75.rds")) {
  model_glm <- glm(high_consumption ~ ., data = train_df, family = binomial)
  saveRDS(model_glm, "../models/glm_binomial/glm_binomial_hh_q75.rds")
} else {
  model_glm <- readRDS("../models/glm_binomial/glm_binomial_hh_q75.rds")
}
```

For each of the 127 households, the 75th percentile of their `log_kWh_total`
is computed and used as that household's personal high-consumption threshold.
A day is labelled 1 if it falls in the top 25% of that household's own
consumption distribution. By construction, every household contributes exactly
25% high-consumption days, so building type and household size are absorbed into
the threshold and do not mechanically drive the outcome. The dataset is
restricted to nine predictors covering weather conditions, building
characteristics, and calendar effects, then split into 80% training and 20% test
sets. The model is fitted with `glm()` using `family = binomial`, which models
the log-odds of high consumption as a linear combination of predictors. The
saved model is reloaded on subsequent knits to avoid refitting.

### Model Results

```{r glm-binomial-performance}
pred_prob <- predict(model_glm, newdata = test_df, type = "response")
pred_class <- ifelse(pred_prob >= 0.5, 1L, 0L)
actual <- test_df$high_consumption

conf_mat <- table(Predicted = pred_class, Actual = actual)
accuracy <- mean(pred_class == actual)

compute_auc <- function(actual, prob) {
  ord <- order(prob, decreasing = TRUE)
  act <- actual[ord]
  n1 <- sum(act == 1)
  n0 <- sum(act == 0)
  tpr <- c(0, cumsum(act == 1) / n1)
  fpr <- c(0, cumsum(act == 0) / n0)
  sum(diff(fpr) * (tpr[-1] + tpr[-length(tpr)]) / 2)
}

glm_auc_val <- compute_auc(actual, pred_prob)
tp <- conf_mat["1", "1"]
tn <- conf_mat["0", "0"]
fp <- conf_mat["1", "0"]
fn <- conf_mat["0", "1"]
precision <- tp / (tp + fp)
recall <- tp / (tp + fn)
f1 <- 2 * precision * recall / (precision + recall)
specificity <- tn / (tn + fp)
```

Coefficients from a binomial GLM are expressed as **odds ratios** (OR =
exp(coefficient)). A value above 1 increases the odds that a day falls in the
top 25% of a household's own consumption, and a value below 1 decreases it,
relative to the reference category.

```{r glm-binomial-coef-table, echo=FALSE}
glm_cs <- coef(summary(model_glm))
glm_tbl <- data.frame(
  Term = rownames(glm_cs),
  Estimate = round(glm_cs[, "Estimate"], 3),
  OR = round(exp(glm_cs[, "Estimate"]), 3),
  CI_Low = round(
    exp(
      glm_cs[, "Estimate"] -
        1.96 * glm_cs[, "Std. Error"]
    ),
    3
  ),
  CI_High = round(
    exp(
      glm_cs[, "Estimate"] +
        1.96 * glm_cs[, "Std. Error"]
    ),
    3
  ),
  p_value = signif(glm_cs[, "Pr(>|z|)"], 3),
  check.names = FALSE
) |>
  filter(Term != "(Intercept)", !grepl("^month", Term)) |>
  arrange(desc(abs(Estimate)))

knitr::kable(
  glm_tbl,
  row.names = FALSE,
  col.names = c("Term", "Estimate", "OR", "CI Low", "CI High", "p-value"),
  caption = paste(
    "GLM Binomial: coefficient estimates and odds ratios.",
    "Month dummies omitted here. All effects shown in the OR plot below."
  )
) %>%
  kable_styling(bootstrap_options = c("striped", "hover"), full_width = TRUE)
```

A striking result of using the per-household threshold is that structural
predictors (building type, heat pump type, living area, and number of residents)
are all statistically non-significant. This is expected because the threshold
is set individually for each household, and these characteristics are already
accounted for in the baseline. The model is therefore answering a different and
more useful question. Given how this household normally behaves, what makes a
specific day unusually high?

The coefficient table reveals several key patterns.

- **Weekend:** The dominant predictor, with OR =
  `r round(exp(coef(model_glm)["is_weekend"]), 2)`. Days falling on a Saturday
  or Sunday have `r round((exp(coef(model_glm)["is_weekend"]) - 1) * 100, 0)`%
  higher odds of being a top-25% consumption day for that household. Households
  spend more time at home on weekends, increasing both heating and appliance
  usage regardless of building type or size.
- **Average temperature:** Each additional degree Celsius reduces the odds by
  `r round((1 - exp(coef(model_glm)["temp_avg"])) * 100, 1)`% (OR =
  `r round(exp(coef(model_glm)["temp_avg"]), 3)`, p < 0.001). Colder days push
  households above their own typical usage level through increased heat pump
  operation.
- **Heating degree days:** Each additional HDD increases the odds by
  `r round((exp(coef(model_glm)["heating_degree_days"]) - 1) * 100, 1)`% (OR =
  `r round(exp(coef(model_glm)["heating_degree_days"]), 3)`, p < 0.001). HDD and 
  temperature are collinear and together capture the thermal load on the
  heat pump. Both are retained because HDD captures heating demand only and is 
  zero on warm days, while temp_avg can detect cooling demand from reversible 
  heat pumps, a pattern HDD alone would miss entirely.
- **Building type, heat pump type, living area, number of residents, floor
  heating:** None of these are statistically significant (all p > 0.05). Once
  the model conditions on a household's own 75th percentile, these structural
  characteristics carry no additional explanatory power for day-to-day
  variation in consumption.
- **Month:** Spring and autumn show the strongest reductions relative to January.
  April and October have the lowest odds (OR
  `r round(exp(coef(model_glm)["month4"]), 2)` and
  `r round(exp(coef(model_glm)["month10"]), 2)` respectively), as these are
  transition months when neither heating nor cooling is needed. Notably, June and July are
  not significantly different from January (p = 0.31 and 0.09), suggesting that
  some households also produce high-consumption days in summer, likely from
  heat pump cooling or increased hot water use. December is likewise
  indistinguishable from January (p = 0.41). The full seasonal pattern is
  visible in the odds ratio plot below.

### Visual Evidence

Each point shows the estimated odds ratio for one predictor, with 95% confidence
intervals. Points to the right of the dashed line (OR > 1) increase the odds of
high consumption, and points to the left decrease them. Blue points are
statistically significant (p < 0.05) and grey points are not. The log scale
ensures that effects of equal magnitude in both directions appear symmetrically.

```{r glm-binomial-or, fig.height=5, fig.width=7, fig.align='center'}
coef_summary <- coef(summary(model_glm))
coef_df <- data.frame(
  term = rownames(coef_summary),
  estimate = coef_summary[, "Estimate"],
  se = coef_summary[, "Std. Error"],
  p_value = coef_summary[, "Pr(>|z|)"]
) |>
  filter(term != "(Intercept)") |>
  mutate(
    odds_ratio = exp(estimate),
    ci_low = exp(estimate - 1.96 * se),
    ci_high = exp(estimate + 1.96 * se),
    significant = p_value < 0.05
  ) |>
  arrange(odds_ratio)

ggplot(
  coef_df,
  aes(x = odds_ratio, y = reorder(term, odds_ratio), color = significant)
) +
  geom_point(size = 2.5) +
  geom_errorbarh(aes(xmin = ci_low, xmax = ci_high), height = 0.3) +
  geom_vline(xintercept = 1, linetype = "dashed", color = clr_neutral) +
  scale_color_manual(
    values = c("FALSE" = clr_neutral, "TRUE" = clr_primary),
    labels = c("n.s.", "p < 0.05")
  ) +
  scale_x_log10() +
  labs(
    title = "GLM Binomial – Odds Ratios with 95% CIs",
    x = "Odds Ratio (log scale)",
    y = NULL,
    color = NULL
  ) +
  report_theme
```

The wide spread of odds ratios shows that calendar effects, particularly month 
dummies, drive the largest deviations from OR = 1, while structural household 
characteristics (building type, heat pump type) cluster near the reference line 
and are statistically non-significant, as expected by the modell setup.

The ROC curve traces the model's sensitivity against its false positive rate as
the classification threshold varies from 0 to 1. The dashed diagonal represents
a model that predicts at random (AUC = 0.5). The further the blue curve bows
toward the top-left corner, the better the model separates the two classes
regardless of threshold choice.

```{r glm-binomial-roc, fig.height=4, fig.width=6, fig.align='center'}
roc_df <- (function(actual, prob) {
  ord <- order(prob, decreasing = TRUE)
  act <- actual[ord]
  n1 <- sum(act == 1)
  n0 <- sum(act == 0)
  data.frame(
    fpr = c(0, cumsum(act == 0) / n0),
    tpr = c(0, cumsum(act == 1) / n1)
  )
})(actual, pred_prob)

ggplot(roc_df, aes(x = fpr, y = tpr)) +
  geom_line(color = clr_primary, linewidth = 1) +
  geom_abline(
    slope = 1,
    intercept = 0,
    linetype = "dashed",
    color = clr_neutral
  ) +
  labs(
    title = "GLM Binomial – ROC Curve",
    subtitle = paste0("AUC = ", round(glm_auc_val, 3)),
    x = "False Positive Rate",
    y = "True Positive Rate"
  ) +
  report_theme
```

With an AUC of `r round(glm_auc_val, 3)`, the model
correctly ranks a randomly drawn high-consumption day above a randomly drawn
low-consumption day `r round(glm_auc_val * 100, 1)`% of the time, which is a
meaningful improvement over chance.

### Model Diagnostics

The confusion matrix shows the model's classifications on the held-out test set.
The model correctly identifies `r round(recall * 100, 1)`% of true high-
consumption days (sensitivity) and `r round(specificity * 100, 1)`% of true
low-consumption days (specificity). Because the per-household 75th percentile
produces a 75/25 class split, a naive classifier that always predicts normal
would achieve around `r round(sum(actual == 0)/length(actual) * 100, 1)`%
accuracy. The model's accuracy of `r round(accuracy * 100, 1)`% is only
`r round((accuracy - sum(actual == 0)/length(actual)) * 100, 1)` percentage
points above this baseline, so AUC is the more informative metric.

```{r glm-binomial-confmat, fig.height=4, fig.width=4.5, fig.align='center'}
conf_df <- data.frame(
  Predicted = factor(c("0", "0", "1", "1"), levels = c("1", "0")),
  Actual = factor(c("0", "1", "0", "1"), levels = c("0", "1")),
  Count = c(tn, fn, fp, tp),
  Label = c("TN", "FN", "FP", "TP")
)
ggplot(conf_df, aes(x = Actual, y = Predicted, fill = Label)) +
  geom_tile(color = "white", linewidth = 1.2) +
  geom_text(
    aes(label = paste0(Label, "\n", Count)),
    size = 5,
    fontface = "bold",
    color = "white"
  ) +
  scale_fill_manual(
    values = c(
      "TN" = clr_primary,
      "TP" = clr_primary,
      "FP" = clr_secondary,
      "FN" = clr_secondary
    )
  ) +
  scale_x_discrete(labels = c("0" = "Low (0)", "1" = "High (1)")) +
  scale_y_discrete(labels = c("0" = "Low (0)", "1" = "High (1)")) +
  labs(
    title = "GLM Binomial – Confusion Matrix",
    subtitle = paste0(
      "Accuracy = ",
      round(accuracy, 3),
      "  |  F1 = ",
      round(f1, 3),
      "  |  AUC = ",
      round(glm_auc_val, 3)
    ),
    x = "Actual class",
    y = "Predicted class",
    fill = "Count"
  ) +
  report_theme +
  theme(legend.position = "none", panel.grid = element_blank())
```

The remaining errors occur mainly on days where consumption is only marginally 
above or below the household's personal threshold and therefore genuinely ambiguous.
The F1 score of 0.685 summarises the balance between precision and recall into a 
single number, useful here because the 75/25 class split means accuracy alone is 
misleading. The AUC of 0.896 measures the model's ability to rank a randomly drawn 
high-consumption day above a randomly drawn low-consumption day, regardless of the 
classification threshold chosen. Together, F1 and AUC confirm that the model 
performs well beyond what the modest accuracy gain over the naive baseline suggests.

### Key Findings

- The model achieves an AUC of `r round(glm_auc_val, 3)` and an accuracy of
  `r round(accuracy * 100, 1)`% on the held-out test set, compared to a naive
  baseline of `r round(sum(actual == 0)/length(actual) * 100, 1)`%. AUC is the
  more informative metric given the 75/25 class split.
- **Weekend is the dominant predictor** (OR =
  `r round(exp(coef(model_glm)["is_weekend"]), 2)`). Occupancy patterns, not
  structural characteristics, are the strongest driver of within-household
  demand spikes.
- **Building type, heat pump type, living area, floor heating, and number of
  residents are all non-significant.** The per-household threshold neutralises
  these structural differences. The model is explaining day-to-day variation
  around each household's own baseline, not variation between households.
- **Cold weather drives high-consumption days** through both the temperature
  coefficient (OR = `r round(exp(coef(model_glm)["temp_avg"]), 3)` per °C) and
  HDD (OR = `r round(exp(coef(model_glm)["heating_degree_days"]), 3)` per HDD).
- **Spring and autumn are low-risk seasons.** April and October show the largest
  reductions relative to January. June and July are not significantly different
  from January, indicating that some households generate high-consumption days
  in summer as well, likely from cooling or increased hot water use.
- **Sensitivity of `r round(recall * 100, 1)`%** means the model correctly
  flags roughly two thirds of actual high-consumption days but also misses one
  in three demand spikes
  entirely. A client using this model for operational alerting should be aware
  of this limitation.

For the client, demand spikes at the household level are predictable from
calendar and weather information alone, without needing to know anything about
the household's physical characteristics — structural variables lose significance
once the model conditions on each household's own baseline. The model is
transparent, interpretable, and provides a well-founded baseline for comparison
with more flexible classification methods.

## Generalised Additive Model (GAM) {#gam}

*Lead: Emanuel Lemma*

### Objective

The Linear Model assumes that each predictor affects the response in a straight
line. For energy consumption, this is often too rigid. The relationship between
heating demand and electricity use may differ between houses and apartments, and
between heat pump types, in ways that cannot be captured by fixed linear slopes.
A **Generalised Additive Model (GAM)** relaxes this assumption by replacing
fixed slopes with flexible smooth functions, while remaining fully interpretable
through its component plots. Here we model `log_kWh_total` using **exactly the
same predictors as the extended Linear Model**, namely heating degree days
(interacted with building type and heat pump type), sunshine hours, humidity,
living area, and the number of residents. The GAM replaces the LM's explicit interaction
terms with `by =` smooths, which allow the shape of the heating-demand curve to
vary freely between groups. This makes the comparison between the two models
direct and fair.

### Model Setup

```{r gam-prep}
df_gam <- heapo |>
  select(
    log_kWh_total,
    heating_degree_days,
    sunshine_hours,
    humidity_avg,
    living_area,
    building_type,
    heatpump_type,
    n_residents
  ) |>
  drop_na()

set.seed(42)
gam_train_idx <- sample(nrow(df_gam), 0.8 * nrow(df_gam))
gam_train <- df_gam[gam_train_idx, ]
gam_test <- df_gam[-gam_train_idx, ]
```

```{r gam-fit, cache=TRUE}
if (!file.exists("../models/gam/gam_model_v2.rds")) {
  model_gam <- mgcv::gam(
    log_kWh_total ~
      s(heating_degree_days, by = building_type) +
      s(heating_degree_days, by = heatpump_type) +
      building_type +
      heatpump_type +
      s(sunshine_hours) +
      s(humidity_avg) +
      s(living_area) +
      n_residents,
    family = gaussian,
    data = gam_train
  )
  saveRDS(model_gam, "../models/gam/gam_model_v2.rds")
} else {
  model_gam <- readRDS("../models/gam/gam_model_v2.rds")
}
```

The model uses `by =` smooths for heating degree days, which allow the heating-
demand curve to take a completely different shape for each building type and heat
pump type, equivalent to the interaction terms in the extended LM but without
constraining the relationship to be linear. All smooths are penalised regression
splines with automatic smoothness selection via restricted maximum likelihood.

$$
\begin{align}
\log(\text{kWh}_i) = \; & \alpha + \beta_1 \text{building type}_i + \beta_2 \text{heatpump type}_i + \beta_3 \text{n residents}_i \\
& + s(\text{HDD}_i \mid \text{building type}_i) + s(\text{HDD}_i \mid \text{heatpump type}_i) \\
& + s(\text{sunshine hours}_i) + s(\text{humidity avg}_i) + s(\text{living area}_i) + \epsilon_i
\end{align}
$$

where $s(\cdot)$ denotes penalised regression splines with automatic smoothness 
selection via Restricted Maximum Likelihood (REML), which balances fit and 
smoothness automatically, and the $by =$ smooths allow the shape of the 
heating-demand curve to vary freely between groups rather than constraining 
it to a fixed linear slope.

### Model Results
The table below summarises the estimated degrees of freedom, F-statistic and 
significance for each smooth term in the model.

```{r gam-smooth-table}
s_gam <- summary(model_gam)
s_tab <- as.data.frame(s_gam$s.table)
s_tab <- cbind(Term = rownames(s_tab), s_tab)
knitr::kable(
  data.frame(
    Term = s_tab$Term,
    EDF = round(s_tab$edf, 1),
    F = round(s_tab$F, 2),
    p.value = formatC(s_tab[["p-value"]], digits = 3, format = "g")
  ),
  caption = "GAM – Smooth Terms",
  col.names = c("Smooth Term", "EDF", "F", "p-value"),
  row.names = FALSE
) %>%
  kable_styling(bootstrap_options = c("striped", "hover"), full_width = TRUE)
```

The smooth terms reveal where non-linearity is present in the data.

- **s(HDD):apartment (edf = 0.8):** The smooth shows a mild positive trend but with a 
  very wide confidence band, reflecting high uncertainty. The non-significance 
  (p = 0.316) is due to this uncertainty rather than the effect being absent.
  For apartments, colder days are associated with higher consumption but the 
  uncertainty is too large to pin down the exact shape of this relationship. 
  This likely reflects higher variability in flat consumption patterns rather 
  than the absence of a heating effect.
- **s(HDD):house (edf = 5.3):** Highly significant (p < 0.001) and strongly
  non-linear. For houses, the response to cold weather is strongly non-linear 
  and highly certain, consumption accelerates disproportionately on the coldest days, 
  meaning a simple straight-line assumption would underestimate peak heating 
  demand.
- **s(HDD):air-source (edf = 1.7):** Air-source heat pumps show a near-linear 
  but slightly curved response to heating demand, consumption rises steadily 
  with cold weather with only a mild bend in the relationship.
- **s(HDD):ground-source (edf = 1.3):** Nearly linear and significant (p =
  0.005). Ground-source heat pumps respond to cold weather in a stable and nearly 
  linear fashion, making their consumption the most predictable of all groups.
- **s(sunshine_hours) (edf = 6.5):** Highly significant (p < 0.001) and
  strongly non-linear. More sunshine generally reduces consumption through 
  passive solar gain, but the relationship is not a simple straight line. 
  The effect levels off at moderate sunshine hours and rises again at very 
  high values, suggesting that on the sunniest days some households switch their
  reversible heat pump to cooling mode, increasing electricity consumption.
- **s(humidity_avg) (edf = 5.3):** Non-linear in shape but not significant
  overall (p = 0.118). Humidity may interact with perceived temperature in ways
  not fully captured here. It is retained to enable a direct and controlled 
  comparison between the LM and GAM on identical predictors. Dropping it did
  not meaningfully change the other coefficients.
- **s(living_area) (edf = 9.0):** Highly significant (p < 0.001) and the most
  non-linear term. Consumption does not scale proportionally with floor area,
  medium-sized homes behave differently from small or very large ones.
  Consumption peaks around 500–600 m² and then declines sharply at very large 
  living areas, which may reflect more energy-efficient construction in larger 
  or newer buildings.

The parametric terms capture the baseline level shifts between groups and are 
summarised in the table below.

```{r gam-parametric-table}
p_tab <- as.data.frame(s_gam$p.table)
p_tab <- cbind(Term = rownames(p_tab), p_tab)
knitr::kable(
  data.frame(
    Term = p_tab$Term,
    Est = round(p_tab$Estimate, 4),
    SE = round(p_tab[["Std. Error"]], 4),
    t.val = round(p_tab[["t value"]], 2),
    p.value = formatC(p_tab[["Pr(>|t|)"]], digits = 3, format = "g")
  ),
  caption = "GAM – Parametric Terms",
  col.names = c("Term", "Estimate", "Std. Error", "t-value", "p-value"),
  row.names = FALSE
) %>%
  kable_styling(bootstrap_options = c("striped", "hover"), full_width = TRUE)
```

While the `by =` smooths allow the heating-demand curve to differ in shape. Converting
the parametric estimates with `(exp(coef) - 1) × 100` gives the percentage
difference relative to the reference level (flat, air-source heat pump).

- **Houses vs. apartments:** The parametric term for `building_typehouse` is
  small and not statistically significant (p = 0.562). Once the model fits a
  separate heating-demand curve per building type, the baseline intercept gap
  disappears. The difference between building types is captured entirely by how
  they respond to cold weather, not by a flat shift.
- **Ground-source vs. air-source heat pump:** Ground-source systems consume
  `r round((1 - exp(coef(model_gam)["heatpump_typeground-source"])) * 100, 1)`%
  less electricity per day, all else equal (p < 0.001). This large reduction
  reflects the inherently higher efficiency of ground-source heat pumps across
  all temperature conditions.
- **Number of residents:** Each additional resident adds
  `r round((exp(coef(model_gam)["n_residents"]) - 1) * 100, 1)`% to daily
  consumption (p < 0.001), confirming that occupancy is a meaningful driver of
  electricity demand.

### Visual Evidence

The component plots below visualise the estimated smooth function for each 
predictor, showing how each variable contributes to the predicted log consumption 
independently of the others.

```{r gam-smooth-plots, fig.height=6, fig.width=9, fig.align='center'}
gratia::draw(
  model_gam,
  residuals = FALSE,
  ci_level = 0.95,
  smooth_col = clr_primary,
  ci_col = clr_primary
) &
  report_theme &
  theme(
    text = element_text(size = 9),
    axis.text = element_text(size = 8),
    strip.text = element_text(size = 9)
  )
```

The seven panels show the estimated smooth function for each term in the model.
The y-axis is the additive contribution to `log_kWh_total`, where positive
values increase predicted consumption and negative values decrease it. The shaded
bands are 95% confidence intervals. The effective degrees of freedom (edf) on
each panel title indicates curvature. An edf of 1 is a straight line and higher
values indicate more complex shapes. The s(HDD):house smooth (edf = 5.3) shows
the most visible curvature, confirming that cold-weather response is the main
source of non-linearity for houses.

To make the by = smooths more interpretable, the plots below show the model's 
predicted daily consumption on the original kWh scale, with all other predictors 
held at their median values.

```{r gam-pred-hdd, fig.height=4, fig.align='center'}
hdd_grid <- expand.grid(
  heating_degree_days = seq(
    min(gam_train$heating_degree_days),
    max(gam_train$heating_degree_days),
    length.out = 200
  ),
  sunshine_hours = median(gam_train$sunshine_hours),
  humidity_avg = median(gam_train$humidity_avg),
  living_area = median(gam_train$living_area),
  n_residents = round(median(gam_train$n_residents)),
  building_type = levels(gam_train$building_type),
  heatpump_type = levels(gam_train$heatpump_type)[1]
)
hdd_pred <- predict(model_gam, newdata = hdd_grid, se.fit = TRUE)
hdd_grid$fit <- hdd_pred$fit
hdd_grid$se <- hdd_pred$se.fit

ggplot(
  hdd_grid,
  aes(
    x = heating_degree_days,
    y = exp(fit),
    color = building_type,
    fill = building_type
  )
) +
  geom_ribbon(
    aes(ymin = exp(fit - 1.96 * se), ymax = exp(fit + 1.96 * se)),
    alpha = 0.15,
    color = NA
  ) +
  geom_line(linewidth = 1) +
  labs(
    title = "GAM – Predicted Daily Consumption by Heating Demand and Building Type",
    subtitle = "Other predictors held at median; air-source heat pump",
    x = "Heating Degree Days",
    y = "Predicted kWh/day",
    color = "Building type",
    fill = "Building type"
  ) +
  scale_colour_manual(values = clr_builds) +
  scale_fill_manual(values = clr_builds) +
  report_theme
```

This plot shows how the GAM's heating-demand curve differs between houses and
apartments. Both groups consume more electricity on colder days, but houses
show a steeper and more curved response, particularly at high HDD values. This
non-linearity, captured by the s(HDD):house smooth with edf = 5.3, would have
been forced into a single linear slope by the LM's interaction term. The wider
confidence band for apartments at extreme cold values reflects lower sample
density there.

```{r gam-pred-heatpump, fig.height=4, fig.align='center'}
hp_grid <- expand.grid(
  heating_degree_days = seq(
    min(gam_train$heating_degree_days),
    max(gam_train$heating_degree_days),
    length.out = 200
  ),
  sunshine_hours = median(gam_train$sunshine_hours),
  humidity_avg = median(gam_train$humidity_avg),
  living_area = median(gam_train$living_area),
  n_residents = round(median(gam_train$n_residents)),
  building_type = levels(gam_train$building_type)[1],
  heatpump_type = levels(gam_train$heatpump_type)
)
hp_pred <- predict(model_gam, newdata = hp_grid, se.fit = TRUE)
hp_grid$fit <- hp_pred$fit
hp_grid$se <- hp_pred$se.fit

ggplot(
  hp_grid,
  aes(
    x = heating_degree_days,
    y = exp(fit),
    color = heatpump_type,
    fill = heatpump_type
  )
) +
  geom_ribbon(
    aes(ymin = exp(fit - 1.96 * se), ymax = exp(fit + 1.96 * se)),
    alpha = 0.15,
    color = NA
  ) +
  geom_line(linewidth = 1) +
  labs(
    title = "GAM – Predicted Daily Consumption by Heating Demand and Heat Pump Type",
    subtitle = "Other predictors held at median; apartment building type",
    x = "Heating Degree Days",
    y = "Predicted kWh/day",
    color = "Heat pump type",
    fill = "Heat pump type"
  ) +
  scale_colour_manual(
    values = c("air-source" = clr_primary, "ground-source" = clr_secondary)
  ) +
  scale_fill_manual(
    values = c("air-source" = clr_primary, "ground-source" = clr_secondary)
  ) +
  report_theme
```

This plot compares the predicted heating-demand curves for air-source and
ground-source heat pumps. Ground-source systems consume consistently less
electricity across the full range of heating demand. The near-linear shape of
both curves (reflecting low edf for these smooths) suggests that heat pump type
shifts the intercept of the heating response more than it changes its curvature.
The stable gap confirms the large parametric coefficient for ground-source heat
pumps identified above.

### Model Diagnostics

The plots below assess how well the model fits the training data and whether 
any systematic patterns remain in the residuals.

```{r gam-obs-vs-fitted, echo=FALSE, fig.height=3.5, fig.width=5.5, out.width="70%", fig.align="center"}
fit_df <- data.frame(
  observed = gam_train$log_kWh_total,
  fitted = fitted(model_gam)
)
ggplot(fit_df, aes(x = fitted, y = observed)) +
  geom_point(alpha = 0.15, size = 0.8, color = clr_primary) +
  geom_abline(
    intercept = 0,
    slope = 1,
    color = clr_secondary,
    linetype = "dashed"
  ) +
  labs(
    title = "GAM – Observed vs. Fitted",
    x = "Fitted log(kWh_total)",
    y = "Observed log(kWh_total)"
  ) +
  report_theme
```

The dashed line is the ideal 1:1 line. A perfectly calibrated model would
scatter all points along it symmetrically. Systematic deviations above or below
indicate that the model consistently over- or under-estimates in certain
consumption ranges. The points follow the 1:1 line reasonably well in the mid-range, 
confirming the model captures the main consumption patterns. A cluster of strongly negative 
observed values around fitted = 2.5 is visible, likely representing near-zero 
consumption days or data anomalies. At low fitted values the cloud sits slightly 
above the reference line, indicating mild underestimation there. Both patterns 
are consistent with a cross-sectional model that cannot account for individual 
household behaviour.

```{r gam-residuals, fig.height=4, fig.align='center'}
resid_df <- data.frame(
  fitted = fitted(model_gam),
  residuals = residuals(model_gam)
)
ggplot(resid_df, aes(x = fitted, y = residuals)) +
  geom_point(alpha = 0.1, size = 0.6, color = clr_primary) +
  geom_hline(yintercept = 0, color = clr_secondary, linetype = "dashed") +
  geom_smooth(
    method = "loess",
    se = FALSE,
    color = clr_secondary,
    linewidth = 0.8
  ) +
  labs(
    title = "GAM – Residuals vs. Fitted",
    x = "Fitted log(kWh_total)",
    y = "Residuals"
  ) +
  report_theme
```

The loess smoother is approximately flat and close to zero across most of the 
fitted range, indicating no major systematic bias. A slight negative dip is 
visible at low fitted values and a small downward drift appears at the high end. 
The vertical streaks of extreme negative residuals around fitted = 2 correspond 
to the same anomalous near-zero consumption days visible in the observed vs 
fitted plot. Residual variance is broadly stable across the fitted range with 
no strong funnel shape, suggesting heteroscedasticity is not a major concern.

```{r gam-performance}
pred_gam <- predict(model_gam, newdata = gam_test)
actual_log <- gam_test$log_kWh_total
actual_kwh <- exp(actual_log)
pred_kwh <- exp(pred_gam)

gam_mae_kwh <- mean(abs(actual_kwh - pred_kwh))
median_kwh <- median(actual_kwh)
mae_pct <- gam_mae_kwh / median_kwh * 100
r2_gam <- 1 -
  sum((actual_log - pred_gam)^2) / sum((actual_log - mean(actual_log))^2)

cat(sprintf(
  "R\u00b2 (test):          %.1f%%\nMAE (kWh/day):     %.1f kWh\nMedian consumption: %.1f kWh/day\nMAE as %% of median: %.1f%%\n",
  r2_gam * 100,
  gam_mae_kwh,
  median_kwh,
  mae_pct
))
```

The most interpretable performance metric here is the **mean absolute error in
kWh**. On average, the model's daily consumption forecasts are off by
**`r round(gam_mae_kwh, 1)` kWh**, against a median household consumption of
**`r round(median_kwh, 1)` kWh/day**. This means a typical prediction error of
about **`r round(mae_pct, 0)`%** of a household's usual daily consumption.
This is expected for a cross-sectional model of this type. 
Weather and building characteristics capture the structural drivers of demand, 
but day-to-day variation driven by individual behaviour, whether someone is 
home, what appliances they use, is simply not observable from the available 
predictors and represents a natural ceiling for this modelling approach.

The **R² of `r round(r2_gam * 100, 1)`%** tells the same story from a different
angle. The eight predictors explain roughly a third of the day-to-day variation
in consumption, and the remaining two-thirds is driven by factors not available
in this dataset. Compared to the extended Linear Model (R² = 27.4%, MAE =
10.96 kWh/day), the GAM achieves a modest improvement. R² rises by 3.6
percentage points and MAE falls by approximately 0.4 kWh/day. This improvement
is real but small, suggesting that the LM's linear approximation is adequate for
most of the predictors in this dataset.

A note on why RMSE was not used. RMSE on the log scale is mathematically correct
but has no intuitive unit that a client can relate to. MAE in kWh is directly
actionable, and R² communicates model quality in plain percentage terms.

### Key Findings

- **Heating demand (HDD)** is non-linear for houses (edf = 5.3, p < 0.001),
  meaning the LM's linear interaction term understates the acceleration in
  consumption during the coldest days.
- **HDD for apartments** shows no significant smooth effect (edf = 0.8,
  p = 0.316). Flat consumption responds to cold weather in a way that is
  well described by a straight line once group-specific curves are allowed.
  The extended LM captures those effects adequately.
- **Ground-source heat pumps** consume
  `r round((1 - exp(coef(model_gam)["heatpump_typeground-source"])) * 100, 1)`%
  less electricity than air-source systems (p < 0.001).
- **Each additional resident** adds
  `r round((exp(coef(model_gam)["n_residents"]) - 1) * 100, 1)`% to daily
  consumption (p < 0.001).
- **Living area** is the most non-linear predictor (edf = 9.0, p < 0.001).
  Consumption does not scale proportionally with floor area.
- The GAM achieves an R² of `r round(r2_gam * 100, 1)`% and a MAE of
  `r round(gam_mae_kwh, 1)` kWh/day, improving on the extended LM by about 0.4
  kWh/day. This modest gain does not justify choosing the GAM over the LM when
  interpretability is the priority. The detailed comparison using 5-fold
  cross-validation is presented in the Model Comparison section.

## Neural Network (NN) {#nn}
*Lead: Raphaël Spagolla*

```{r import-NN, echo=FALSE, message=FALSE, warning=FALSE}
mod_nn <- readRDS("../models/neural_network/mod_nn_nnet.rds")
test_nn <- readRDS("../models/neural_network/splits/test_nn.rds")
pred_prob_1 <- readRDS("../models/neural_network/splits/nn_pred_prob.rds")
```

### Objective

The aim of this analysis is to predict whether an observation corresponds to a
high-consumption event (high_consumption), a binary outcome taking values 0 or
1. This is a classification problem, where the focus is on accurately
distinguishing between high- and low-consumption cases.

From a practical perspective, reliable classification of high-consumption events
can support decision-making processes such as resource allocation, cost
management, or demand planning. Early identification of such events allows for
more efficient operational responses.

A neural network is employed to capture potentially complex and non-linear
relationships between consumption and predictors, including weather conditions,
building characteristics, and household composition. Model performance is
evaluated based on predictive accuracy and classification metrics derived from
test data.

### Data Preparation

The dataset consists of 99,838 observations and includes variables related to
weather conditions, building characteristics, and household composition. The
response variable, *high_consumption*, is binary, defined using each household's own 75th percentile of `log_kWh_total` as the threshold — consistent with the GLM Binomial and SVM. By construction, exactly 25% of each household's days are classified as high-consumption.

To ensure valid model evaluation, the data is split into training (80%) and test
(20%) sets using a stratified sampling approach. This preserves the class
distribution in both subsets and avoids biased performance estimates that could
arise from imbalanced splits.

Categorical variables are transformed into dummy variables using one-hot
encoding, allowing them to be used as inputs in the neural network. In addition,
all predictors are scaled based on the training data. This step is important for
neural networks, as it improves numerical stability and facilitates efficient
training. The same transformations are then applied to the test set to ensure
consistency and prevent data leakage.

### Model setup

<details>
<summary>Show / hide result</summary>
```{r nn-modell, echo=TRUE}
cat(deparse(formula(mod_nn)), "\n\n")
mod_nn$call
```
</details>


The neural network is specified with an architecture of (9–25–1), consisting of
nine input nodes, one hidden layer with 25 neurons, and a single output
neuron for binary classification. This relatively simple structure is chosen to
balance model flexibility with the risk of overfitting, allowing the network to
capture non-linear relationships without introducing unnecessary complexity.

A non-linear activation function is applied in the hidden layer, enabling the
model to learn complex patterns in the data. The output layer produces predicted
probabilities for the high-consumption class.

To improve generalisation, weight decay (set to 0.05) is included as a
regularisation mechanism. This penalises large weights and helps reduce model
variance. The network is trained using the nnet package with a sufficient number
of iterations to ensure convergence.

Overall, the chosen specification reflects a trade-off between predictive
performance and model stability, favouring a parsimonious architecture that
performs reliably on unseen data.

### Model Results
<details>
<summary>Show / hide result</summary>
```{r nn-results, echo=TRUE}
truth <- test_nn$high_consumption
pred_class <- factor(
  ifelse(pred_prob_1 >= 0.5, "1", "0"),
  levels = levels(truth)
)

cm <- table(Predicted = pred_class, Actual = truth)
acc <- mean(pred_class == truth)
tp <- cm["1", "1"]
tn <- cm["0", "0"]
fp <- cm["1", "0"]
fn <- cm["0", "1"]
sensitivity <- tp / (tp + fn)
specificity <- tn / (tn + fp)

# ROC object (used by plots below)
pred_roc <- ROCR::prediction(pred_prob_1, truth)
perf_roc <- ROCR::performance(pred_roc, "tpr", "fpr")
nn_auc_val <- round(
  as.numeric(ROCR::performance(pred_roc, "auc")@y.values[[1]]),
  3
)

knitr::kable(
  data.frame(
    Model = "Neural Network",
    Accuracy = round(acc, 3),
    Sensitivity = round(sensitivity, 3),
    Specificity = round(specificity, 3),
    AUC = nn_auc_val
  ),
  caption = "Neural Network Performance Metrics"
) %>%
  kable_styling(bootstrap_options = c("striped", "hover"), full_width = TRUE)
```
</details>


The neural network achieves an accuracy of approximately `r round(acc, 2)`. Because the binary
target is defined using the 75th-percentile threshold, roughly 75% of observations
belong to the low-consumption class; a naive classifier that always predicts "low"
would therefore achieve a baseline accuracy of $\approx 0.75$. The model's accuracy
must be assessed against this higher bar rather than a 50/50 split. The AUC of
`r nn_auc_val` provides a threshold-independent measure and confirms discriminative
ability beyond the class-imbalance baseline.

Specificity (`r round(specificity, 3)`) exceeds sensitivity (`r round(sensitivity, 3)`), indicating that the model is
more effective at identifying low-consumption days than high-consumption
days. This is expected given the 75/25 class imbalance: with three times as many
normal days as high-consumption days, the model has more signal for the majority
class. The lower sensitivity reflects a higher rate of missed high-consumption
events, which is the more costly error in capacity-planning contexts.

Overall, the confusion matrix shows that the majority of observations are
correctly classified, with most predictions concentrated along the diagonal.
However, misclassifications occur in both directions, indicating that some
observations share similar characteristics across classes.

These results indicate good overall classification performance, although some
variability remains unexplained, likely due to overlap in predictor patterns
between the two classes.

### Visual Evidence

The *predicted probability plot* illustrates how the neural network separates
the two classes. Observations belonging to the high-consumption class tend to
have higher predicted probabilities, while low-consumption cases are
concentrated at lower values. However, noticeable overlap in the mid-range
indicates that some observations cannot be clearly distinguished, which explains
the presence of classification errors.

```{r nn-pred-prob, echo=TRUE, fig.height=4, fig.width=6, fig.align='center'}
nn_prob_df <- data.frame(
  index = seq_along(pred_prob_1),
  probability = as.numeric(pred_prob_1),
  actual = as.character(truth)
)

ggplot(nn_prob_df, aes(x = index, y = probability, color = actual)) +
  geom_point(alpha = 0.4, size = 0.4) +
  geom_hline(yintercept = 0.5, linetype = "dashed", color = "grey50") +
  scale_color_manual(
    values = c("0" = clr_primary, "1" = clr_secondary),
    labels = c("0" = "Actual Low (0)", "1" = "Actual High (1)")
  ) +
  labs(
    title = "Neural Network – Predicted Probabilities",
    x = "Observation",
    y = "Predicted Probability",
    color = NULL
  ) +
  report_theme +
  theme(legend.position = "bottom")
```

The *confusion matrix* provides a direct summary of classification performance.
Most observations lie on the diagonal, confirming a high number of correct
predictions for both classes. Nevertheless, errors occur in both directions,
highlighting that class separation is not complete.

```{r nn-confmat, echo=TRUE, fig.height=4, fig.width=4.5, fig.align='center'}
nn_df <- data.frame(
  Predicted = factor(c("0", "0", "1", "1"), levels = c("1", "0")),
  Actual = factor(c("0", "1", "0", "1"), levels = c("0", "1")),
  Count = c(cm["0", "0"], cm["0", "1"], cm["1", "0"], cm["1", "1"]),
  Label = c("TN", "FN", "FP", "TP")
)

ggplot(nn_df, aes(x = Actual, y = Predicted, fill = Count)) +
  geom_tile(color = "white", linewidth = 1.2) +
  geom_text(
    aes(label = paste0(Label, "\n", Count)),
    size = 5,
    fontface = "bold",
    color = "white"
  ) +
  scale_fill_gradient(low = clr_primary, high = clr_secondary) +
  scale_x_discrete(labels = c("0" = "Low (0)", "1" = "High (1)")) +
  scale_y_discrete(labels = c("0" = "Low (0)", "1" = "High (1)")) +
  labs(
    title = "Neural Network – Confusion Matrix",
    subtitle = paste0(
      "Accuracy = ",
      round(acc, 3),
      "  |  F1 = ",
      round(
        2 *
          (tp / (tp + fp)) *
          (tp / (tp + fn)) /
          ((tp / (tp + fp)) + (tp / (tp + fn))),
        3
      ),
      "  |  AUC = ",
      nn_auc_val
    ),
    x = "Actual class",
    y = "Predicted class"
  ) +
  report_theme +
  theme(legend.position = "none", panel.grid = element_blank())
```

The *ROC curve* further evaluates model performance across all possible
classification thresholds. The curve lies clearly above the diagonal reference
line, indicating strong discriminative ability and performance well above random
classification. The steep initial increase suggests that high true positive
rates can be achieved with relatively low false positive rates, while the
flattening at higher values reflects diminishing gains.

```{r nn_roc, echo=TRUE, fig.height=4, fig.width=6, fig.align='center'}
roc_df_nn <- data.frame(
  fpr = perf_roc@x.values[[1]],
  tpr = perf_roc@y.values[[1]]
)

ggplot(roc_df_nn, aes(x = fpr, y = tpr)) +
  geom_line(color = clr_primary, linewidth = 1) +
  geom_abline(
    slope = 1,
    intercept = 0,
    linetype = "dashed",
    color = clr_neutral
  ) +
  labs(
    title = "Neural Network – ROC Curve",
    subtitle = paste0("AUC = ", nn_auc_val),
    x = "False Positive Rate",
    y = "True Positive Rate"
  ) +
  report_theme
```

### Model Diagnostics

Neural network diagnostics focus on predictive performance rather than
distributional assumptions. The chosen architecture (9–25–1) provides a balance
between flexibility and overfitting, allowing the model to capture non-linear
relationships while maintaining stable generalisation.

The test accuracy of approximately `r round(acc * 100, 1)`% indicates reasonable generalisation to
unseen data. The absence of a substantial gap between training and test
performance suggests that overfitting is limited, although some dependence on
the specific train–test split may remain.

The ROC curve confirms consistent performance across classification thresholds
and a clear improvement over random classification. However, the observed
overlap in predicted probabilities indicates that the available predictors do
not fully separate the two classes. This suggests that model performance is
constrained more by the information content of the data than by the chosen model
structure.

Alternative network configurations were explored to assess the impact of model
complexity. Increasing the number of hidden neurons led to slight improvements
in training performance but did not translate into meaningful gains in test
accuracy, indicating potential overfitting. Simpler architectures achieved
comparable results with greater stability, supporting the choice of the 9–25–1
specification.

### Coefficient Interpretation

Neural network weights are not directly interpretable as coefficients. Because
inputs pass through non-linear transformations across multiple layers, no single
weight carries an isolated, meaningful effect. Model performance metrics (AUC,
accuracy) and the cross-validation comparison in the Model Comparison section
serve as the practical evaluation instead.

### Key Findings

* The neural network achieves solid predictive performance and successfully
  captures non-linear relationships in the data. Accuracy is approximately
  `r round(acc * 100, 1)`%, above the $\approx 75$% naive baseline arising from
  the 75/25 class split, though the margin is modest.

* The model is more effective at identifying low-consumption days than high-
  consumption days, as reflected in higher specificity compared to sensitivity.
  This is expected given the 75/25 class imbalance.

* The confusion matrix and ROC curve confirm good overall classification
  ability, although not perfect.

* Overlap in predicted probabilities indicates that some observations cannot be
  clearly separated, limiting maximum achievable performance.

* Increasing model complexity does not lead to substantial improvements,
  suggesting that performance is primarily constrained by the available features
  rather than the model itself.

From a practical perspective, the model can serve as a useful decision-support
tool for identifying high-consumption events, although predictions should be
interpreted with some caution due to remaining classification uncertainty.
Further improvements are more likely to come from enhanced feature engineering
or the inclusion of additional relevant variables. Cross-validation results for
the Neural Network are reported in the Model Comparison section.

## Support Vector Machine (SVM) {#svm}

*Lead: Tharrmeehan Krishnathasan*

```{r import-svm, echo=TRUE, message=FALSE, warning=FALSE}
hh_q75_svm <- heapo |>
  group_by(Household_ID) |>
  summarise(
    hh_q75 = quantile(log_kWh_total, 0.75, na.rm = TRUE),
    .groups = "drop"
  )

heapo_svm_data <- heapo |>
  left_join(hh_q75_svm, by = "Household_ID") |>
  mutate(
    high_consumption = as.factor(
      ifelse(log_kWh_total >= hh_q75, "high", "normal")
    )
  ) |>
  select(
    high_consumption,
    heating_degree_days,
    sunshine_hours,
    humidity_avg,
    living_area,
    n_residents,
    building_type,
    heatpump_type
  ) |>
  na.omit()

set.seed(42)
train_idx <- createDataPartition(
  heapo_svm_data$high_consumption,
  p = 0.8,
  list = FALSE
)
train_data <- heapo_svm_data[train_idx, ]
test_data <- heapo_svm_data[-train_idx, ]
test_x <- test_data |> select(-high_consumption)
test_y <- test_data$high_consumption

# Baseline SVM: C=1, sigma estimated from training data via sigest.
# Cached to heapo_svm_baseline_hh_q75.rds so subsequent knits skip refit.
if (!file.exists("../models/svm/heapo_svm_baseline_hh_q75.rds")) {
  svm_baseline <- kernlab::ksvm(
    high_consumption ~ .,
    data = train_data,
    kernel = "rbfdot",
    C = 1,
    kpar = "automatic",
    scaled = TRUE
  )
  saveRDS(svm_baseline, "../models/svm/heapo_svm_baseline_hh_q75.rds")
} else {
  svm_baseline <- readRDS("../models/svm/heapo_svm_baseline_hh_q75.rds")
}

# Tuned SVM: 5-fold CV grid search over C and sigma, same grid as original.
# Cached to heapo_svm_hh_q75.rds so subsequent knits skip refit.
if (!file.exists("../models/svm/heapo_svm_hh_q75.rds")) {
  tune_grid <- expand.grid(C = c(1, 10, 100), sigma = c(0.01, 0.1, 1 / 7))
  ctrl <- trainControl(method = "cv", number = 5)
  svm_caret <- train(
    high_consumption ~ .,
    data = train_data,
    method = "svmRadial",
    tuneGrid = tune_grid,
    trControl = ctrl,
    preProcess = c("center", "scale")
  )
  saveRDS(svm_caret, "../models/svm/heapo_svm_hh_q75.rds")
} else {
  svm_caret <- readRDS("../models/svm/heapo_svm_hh_q75.rds")
}

pred_best <- predict(svm_caret, test_x)
cm_best <- confusionMatrix(pred_best, test_y, positive = "high")

pred_baseline <- kernlab::predict(svm_baseline, test_x)
cm_baseline <- confusionMatrix(pred_baseline, test_y, positive = "high")

best_C <- svm_caret$bestTune$C
best_sigma <- round(svm_caret$bestTune$sigma, 4)

# 5-fold CV results (pre-computed by R/04_model_comparison_cv_classification.R)
if (file.exists("../models/cv_class_results.rds")) {
  cv_class <- readRDS("../models/cv_class_results.rds")
  svm_cv_row <- cv_class[grepl("SVM", cv_class$Model), ]
  svm_cv_auc <- svm_cv_row[["CV AUC"]]
  svm_cv_acc <- svm_cv_row[["CV Accuracy"]]
} else {
  svm_cv_auc <- NA
  svm_cv_acc <- NA
}
```

### Objective

This section classifies each household-day as *high consumption* (top 25% of that household's days) or *normal*, a binary problem with practical relevance: knowing which days are likely to be high-load in advance lets grid operators and energy providers plan capacity.

An SVM is used because it can find non-linear decision boundaries via kernel functions without requiring distributional assumptions about the predictors. Performance is evaluated on a held-out test set using accuracy, sensitivity, specificity, and F1 score.

### Data Preparation

`high_consumption` labels a day as *high* if it falls in the top 25% of that household's own `log_kWh_total` distribution. For each of the `r length(unique(heapo$Household_ID))` households, the 75th percentile is computed separately, so every household contributes exactly 25% high-consumption days by construction.

After removing missing values, the dataset has 93,686 observations: 23,440 high-consumption days (25.0%) and 70,246 normal days (75.0%). `log_kWh_total` is excluded to prevent data leakage; the SVM must identify high-load days from weather and household characteristics alone.

The data are split 80/20 into training (74,949 rows) and test (18,737 rows) using stratified sampling to keep the class ratio consistent. All numeric predictors are standardised before training, as SVMs are sensitive to feature scale.

<details>
<summary>Show / hide table</summary>
```{r svm-class-balance, echo=TRUE}
prop.table(table(train_data$high_consumption)) %>%
  as.data.frame() %>%
  rename(Class = Var1, Proportion = Freq) %>%
  mutate(Proportion = round(Proportion, 3)) %>%
  knitr::kable(
    caption = "Class balance in the training set"
  ) %>%
  kable_styling(bootstrap_options = c("striped", "hover"), full_width = TRUE)
```
</details>

### Model Setup

Both models use a radial basis function (RBF) kernel, which maps data into a higher-dimensional space to find non-linear decision boundaries. This suits energy data where weather and household effects interact in ways a linear boundary cannot capture.

**Baseline SVM:** C = 1, sigma estimated from training data. This provides a reference for the tuned model.

**Tuned SVM:** 5-fold cross-validation over a grid of `C` ∈ {1, 10, 100} and `sigma` ∈ {0.01, 0.1, 1/7}. `C` controls the misclassification penalty (higher = tighter fit); `sigma` is the inverse bandwidth of the RBF kernel (higher = more local boundaries). The combination with the highest CV accuracy is selected.

The best parameters were C = 100, sigma = 0.1429 (= 1/7), achieving CV accuracy of 0.8304 (Kappa = 0.4859) across the 3×3 grid.

### Model Results

<details>
<summary>Show / hide table</summary>
```{r svm-eval-table, echo=TRUE}
extract_metrics <- function(cm, model_name) {
  tibble(
    Model = model_name,
    Accuracy = round(cm$overall["Accuracy"], 4),
    Sensitivity = round(cm$byClass["Sensitivity"], 4),
    Specificity = round(cm$byClass["Specificity"], 4),
    `F1 Score` = round(cm$byClass["F1"], 4)
  )
}

bind_rows(
  extract_metrics(cm_baseline, "Baseline SVM (C=1, default sigma)"),
  extract_metrics(
    cm_best,
    paste0("Tuned SVM (C=", best_C, ", sigma=", best_sigma, ")")
  )
) %>%
  knitr::kable(
    caption = "SVM model comparison on test set"
  ) %>%
  kable_styling(bootstrap_options = c("striped", "hover"), full_width = TRUE)
```
</details>
<details>
<summary>Show / hide table</summary>
```{r svm-cv-table, echo=TRUE}
if (!is.na(svm_cv_auc)) {
  knitr::kable(
    data.frame(
      Model = paste0("Tuned SVM (C=", best_C, ", σ=", best_sigma, ")"),
      `CV AUC` = round(svm_cv_auc, 3),
      `CV Accuracy` = round(svm_cv_acc, 3),
      check.names = FALSE
    ),
    caption = "Tuned SVM — 5-fold cross-validation performance (per-household 75th percentile threshold)"
  ) %>%
    kable_styling(bootstrap_options = c("striped", "hover"), full_width = TRUE)
} else {
  cat(
    "*CV results not available. Run R/04_model_comparison_cv_classification.R first.*"
  )
}
```
</details>

The tuned SVM achieves a five-fold CV AUC of **`r round(svm_cv_auc, 3)`** and accuracy of **`r round(svm_cv_acc * 100, 1)`%**. The 80/20 split below adds sensitivity, specificity, and F1.

The tuned SVM (C = `r best_C`, sigma = `r best_sigma`) reaches **`r round(cm_best$overall["Accuracy"] * 100, 1)`%** accuracy, compared to `r round(cm_baseline$overall["Accuracy"] * 100, 1)`% for the baseline. The improvement is modest: a naive all-*normal* classifier would already score `r round(table(test_y)["normal"] / length(test_y) * 100, 1)`%, leaving a true gain of only `r round((cm_best$overall["Accuracy"] - table(test_y)["normal"] / length(test_y)) * 100, 0)` percentage points.

The more informative metrics are:

**Sensitivity (`r round(cm_best$byClass["Sensitivity"], 4)`):** the model correctly identifies `r round(cm_best$byClass["Sensitivity"] * 100, 1)`% of actual high-consumption days. Over half are missed, a meaningful limitation in early-warning contexts where false negatives are costly.

**Specificity (`r round(cm_best$byClass["Specificity"], 4)`):** normal days are classified correctly `r round(cm_best$byClass["Specificity"] * 100, 1)`% of the time. The model is conservative, flagging fewer high-consumption days than actually occur (`r sum(cm_best$table["high", ])` predicted high vs. `r sum(cm_best$table[, "high"])` actual high in the test set).

**F1 Score (`r round(cm_best$byClass["F1"], 4)`):** the harmonic mean of precision and recall for the high class reflects the imbalance between high specificity and low sensitivity.

**Kappa (`r round(cm_best$overall["Kappa"], 4)`):** moderate agreement beyond chance, improving on the baseline (`r round(cm_baseline$overall["Kappa"], 4)`) but below the 0.6 threshold for *substantial* agreement.

### Visual Evidence

The confusion matrix heatmap shows the distribution of predictions versus actual labels. Correct predictions fall on the diagonal; off-diagonal cells are misclassifications.

```{r svm-conf-matrix-plot, echo=TRUE, fig.height=4, fig.width=4.5, fig.align='center'}
cm_mat <- cm_best$table

svm_df <- data.frame(
  Predicted = factor(
    c("normal", "normal", "high", "high"),
    levels = c("high", "normal")
  ),
  Actual = factor(
    c("normal", "high", "normal", "high"),
    levels = c("normal", "high")
  ),
  Count = c(
    cm_mat["normal", "normal"],
    cm_mat["normal", "high"],
    cm_mat["high", "normal"],
    cm_mat["high", "high"]
  ),
  Label = c("TN", "FN", "FP", "TP")
)

ggplot(svm_df, aes(x = Actual, y = Predicted, fill = Count)) +
  geom_tile(color = "white", linewidth = 1.2) +
  geom_text(
    aes(label = paste0(Label, "\n", Count)),
    size = 5,
    fontface = "bold",
    color = "white"
  ) +
  scale_fill_gradient(low = clr_primary, high = clr_secondary) +
  scale_x_discrete(labels = c("normal" = "Low (0)", "high" = "High (1)")) +
  scale_y_discrete(labels = c("normal" = "Low (0)", "high" = "High (1)")) +
  labs(
    title = "SVM – Confusion Matrix",
    subtitle = paste0(
      "Accuracy = ",
      round(cm_best$overall["Accuracy"], 3),
      "  |  F1 = ",
      round(cm_best$byClass["F1"], 3),
      "  |  AUC = ",
      round(svm_cv_auc, 3)
    ),
    x = "Actual class",
    y = "Predicted class"
  ) +
  report_theme +
  theme(legend.position = "none", panel.grid = element_blank())
```

The bar chart compares sensitivity and specificity for the baseline and tuned models, showing how tuning shifts the balance between detecting high-consumption days and avoiding false alarms.

```{r svm-metric-comparison, echo=TRUE, fig.height=3.5, fig.width=5, out.width="48%", fig.show="hold", fig.align='center'}
metrics_df <- bind_rows(
  extract_metrics(cm_baseline, "Baseline"),
  extract_metrics(cm_best, "Tuned")
) %>%
  select(Model, Sensitivity, Specificity) %>%
  tidyr::pivot_longer(
    cols = c(Sensitivity, Specificity),
    names_to = "Metric",
    values_to = "Value"
  )

ggplot(metrics_df, aes(x = Metric, y = Value, fill = Model)) +
  geom_col(position = "dodge") +
  scale_fill_manual(
    values = c("Baseline" = clr_neutral, "Tuned" = clr_primary)
  ) +
  scale_y_continuous(limits = c(0, 1)) +
  report_theme +
  labs(
    title = "Sensitivity and Specificity: Baseline vs. Tuned SVM",
    x = NULL,
    y = "Value",
    fill = "Model"
  )
```

### Model Diagnostics

Unlike regression models, SVM diagnostics focus on generalisation performance and modelling choices rather than distributional assumptions.

**CV tuning results:** Across all 9 combinations, CV accuracy ranged from 0.8071 (C=10, sigma=0.01) to 0.8304 (C=100, sigma=0.1429). The top four combinations all involve sigma ≥ 0.1, suggesting a moderately local kernel works best. The gain from increasing C from 10 to 100 at sigma=0.1429 is small (0.8268 → 0.8304), with diminishing returns beyond that.

**Class separability:** The available predictors do not reliably separate the two classes. High and normal consumption days share substantial overlap in the feature space, which no kernel choice can resolve.

**Tuning stability:** The best combination (C=`r best_C`, sigma=`r best_sigma`) sits at the top-right corner of the grid. The gain from C=10 to C=100 at sigma=0.1429 is small (0.8268 → 0.8304), pointing to the predictor set as the limiting factor rather than model configuration.

### Key Findings

**The SVM achieves reasonable accuracy but low sensitivity:** At `r round(cm_best$overall["Accuracy"] * 100, 1)`% accuracy, the model scores roughly 8 percentage points above the naive majority-class baseline (`r round(table(test_y)["normal"] / length(test_y) * 100, 1)`%). Sensitivity of `r round(cm_best$byClass["Sensitivity"] * 100, 1)`% is the bigger concern: more than half of actual high-consumption days are missed, while specificity of `r round(cm_best$byClass["Specificity"] * 100, 1)`% shows the model rarely raises false alarms on normal days. This makes it too conservative on its own for capacity planning or demand-response triggering.

**Tuning yields real but limited improvement:** Moving from default parameters to C=`r best_C`, sigma=`r best_sigma` improves sensitivity from `r round(cm_baseline$byClass["Sensitivity"] * 100, 1)`% to `r round(cm_best$byClass["Sensitivity"] * 100, 1)`% and accuracy from `r round(cm_baseline$overall["Accuracy"] * 100, 1)`% to `r round(cm_best$overall["Accuracy"] * 100, 1)`%. The gains are consistent but small; further tuning will not resolve the core problem.

**The predictor set is the bottleneck:** Weather conditions and static household characteristics capture only part of what drives high-consumption days. High and normal consumption days overlap substantially in the feature space, and no kernel or regularisation choice can separate what is genuinely mixed in the data. Richer features would help: lagged consumption, occupancy patterns, or appliance-level data.

**Threshold adjustment is a practical alternative:** Lowering the flagging threshold would recover more true positives at the cost of more false alarms, which may be acceptable in a probabilistic forecasting pipeline. At the default threshold, the model should not be the sole basis for high-consumption day decisions.

# Model Comparison {#comparison}

The six models do not all share the same response variable, so a single
unified comparison table is not possible. Instead, models are grouped by
what they predict. Regression models (LM and GAM) are evaluated on
`log_kWh_total` using five-fold cross-validation mean absolute error in kWh
and R². Classification models (GLM Binomial, Neural Network, SVM) are
evaluated on `high_consumption` using five-fold CV AUC and accuracy. The
GLM Poisson targets a monthly count at a different aggregation level and is
assessed separately. In all cases, the same random fold assignment is used
across models within each group so that fold-to-fold variation cancels out
and differences reflect only model quality.

## Regression Models: LM vs GAM

```{r reg-comparison-load, echo=FALSE}
if (file.exists("../models/cv_reg_results.rds")) {
  reg_summary <- readRDS("../models/cv_reg_results.rds")
} else {
  reg_summary <- NULL
}
```

The table below reports five-fold cross-validation performance for the two
regression models fitted to `log_kWh_total`. MAE is computed on the original
kWh scale for interpretability. R² is computed on the log scale.

```{r reg-comparison-table, echo=FALSE}
if (!is.null(reg_summary)) {
  knitr::kable(
    reg_summary,
    caption = paste(
      "Five-fold CV performance: regression models on log_kWh_total.",
      "MAE is on the kWh scale; R² is on the log scale."
    )
  ) %>%
    kable_styling(bootstrap_options = c("striped", "hover"), full_width = TRUE)
} else {
  cat("*Table not available. Run R/04_model_comparison_cv.R first.*")
}
```

```{r reg-comparison-plot, echo=FALSE, fig.height=3.5, fig.align='center'}
if (!is.null(reg_summary)) {
  reg_long <- tidyr::pivot_longer(
    reg_summary,
    cols = -Model,
    names_to = "Metric",
    values_to = "Value"
  )

  print(
    ggplot(reg_long, aes(x = Model, y = Value, fill = Model)) +
      geom_col(width = 0.5, show.legend = FALSE) +
      scale_fill_manual(values = c(clr_primary, clr_secondary)) +
      facet_wrap(~Metric, scales = "free_y") +
      report_theme +
      labs(
        title = "5-fold CV: LM vs GAM",
        x = NULL,
        y = NULL
      )
  )
}
```

Both models use **exactly the same predictor set**: heating degree days
(interacted with building type and heat pump type), sunshine hours, humidity,
living area, and number of residents. The only difference is functional form:
the LM uses linear terms and explicit interaction dummies, while the GAM
replaces these with `by =` smooths. This makes the comparison controlled and
directly attributable to model flexibility rather than predictor choice.

The GAM achieves a CV MAE of
`r reg_summary[grepl("GAM", reg_summary$Model), "CV MAE (kWh/day)"]` kWh/day
and an R² of `r reg_summary[grepl("GAM", reg_summary$Model), "CV R2"]`,
compared to
`r reg_summary[grepl("Linear", reg_summary$Model), "CV MAE (kWh/day)"]` kWh/day
and R² of `r reg_summary[grepl("Linear", reg_summary$Model), "CV R2"]` for the
extended LM. The GAM reduces prediction error by
`r round(reg_summary[grepl("Linear", reg_summary$Model), "CV MAE (kWh/day)"] - reg_summary[grepl("GAM", reg_summary$Model), "CV MAE (kWh/day)"], 2)`
kWh/day, a roughly 3.6% improvement, and explains 3.6 additional percentage
points of variance. The improvement is real but modest. As the GAM smooth plots
showed, the main non-linearity is in how houses respond to extreme cold (edf =
5.3 for s(HDD):house), but this effect does not translate into a large
out-of-sample accuracy gain. This suggests that household-level heterogeneity,
not functional form, is the primary source of unexplained variance.

For a client who prioritises interpretability, the extended LM is a reasonable
choice: its coefficients are directly actionable and its predictions are nearly
as accurate. For a client who prioritises predictive performance, the GAM is
preferred, but the marginal gain of 0.40 kWh/day should be weighed against the
added complexity of smooth terms and the loss of a single interpretable
coefficient per predictor.

## Classification Models: GLM Binomial vs SVM vs Neural Network

```{r class-comparison-load, echo=FALSE}
if (file.exists("../models/cv_class_results.rds")) {
  class_summary <- readRDS("../models/cv_class_results.rds")
} else {
  class_summary <- NULL
}
```

The table below compares all three classification models. All three use each 
household's own 75th percentile as the high-consumption threshold, making 
results directly comparable. CV results are available for all three models.

```{r class-comparison-table, echo=FALSE}
if (!is.null(class_summary)) {
  knitr::kable(
    class_summary,
    caption = paste(
      "Classification model performance.",
      "All models: 5-fold CV, per-household 75th percentile threshold."
    )
  ) %>%
    kable_styling(bootstrap_options = c("striped", "hover"), full_width = TRUE)
} else {
  cat(
    "*CV table not available. Run R/04_model_comparison_cv_classification.R then R/3.5_neural_network.R first.*"
  )
}
```

```{r class-comparison-plot, echo=FALSE, fig.height=3.5, fig.align='center'}
if (!is.null(class_summary)) {
  plot_data <- tidyr::pivot_longer(
    class_summary,
    cols = -Model,
    names_to = "Metric",
    values_to = "Value"
  )
  print(
    ggplot(plot_data, aes(x = Model, y = Value, fill = Model)) +
      geom_col(width = 0.5, show.legend = FALSE) +
      scale_fill_manual(values = c(clr_primary, clr_neutral, clr_secondary)) +
      facet_wrap(~Metric, scales = "free_y") +
      report_theme +
      labs(
        title = "Classification model comparison",
        subtitle = "All models: per-household 75th percentile threshold",
        x = NULL,
        y = NULL
      )
  )
}
```

All three models share the same per-household 75th percentile threshold, making them directly comparable. The GLM
Binomial achieves a higher AUC (`r round(class_summary[grepl("GLM", class_summary$Model), "CV AUC"], 3)` vs `r round(class_summary[grepl("SVM", class_summary$Model), "CV AUC"], 3)`) while accuracy is essentially tied (`r round(class_summary[grepl("GLM", class_summary$Model), "CV Accuracy"] * 100, 1)`% vs `r round(class_summary[grepl("SVM", class_summary$Model), "CV Accuracy"] * 100, 1)`%). The AUC gap reflects the GLM's stronger overall discrimination across all classification thresholds. Both models correctly identify roughly four out of five household-days. The GLM Binomial is the preferred choice: it matches the SVM on accuracy, outperforms it on AUC, and provides explicit odds ratios for each predictor. For a client who needs to explain predictions (for example, when flagging a household for an energy audit), this interpretability advantage is directly actionable.

## GLM Poisson (Standalone)

The GLM Poisson models `n_high_days` at the household–month level and
cannot be placed in either comparison table. It is evaluated separately
via five-fold CV RMSE on the count scale.

```{r poisson-cv, echo=TRUE, cache=TRUE}
set.seed(42)

# Extract formula cleanly — detach from model environment to avoid memory bleed
cv_form <- as.formula(
  deparse(formula(mod_pois)),
  env = baseenv()
)

n <- nrow(dat_count)
folds <- sample(rep(1:5, length.out = n))
rmse_folds <- numeric(5)

for (k in 1:5) {
  train_k <- dat_count[folds != k, ]
  test_k <- dat_count[folds == k, ]

  fit_k <- glm(
    cv_form,
    family = poisson(link = "log"),
    offset = log(n_days),
    data = train_k,
    model = FALSE,
    x = FALSE,
    y = FALSE
  )

  pred_k <- predict(fit_k, newdata = test_k, type = "response")
  rmse_folds[k] <- sqrt(mean((test_k$n_high_days - pred_k)^2))
  rm(fit_k, train_k, test_k)
  gc()
}

pois_cv_rmse <- round(mean(rmse_folds), 3)
pois_cv_sd <- round(sd(rmse_folds), 3)
```

The five-fold CV RMSE is `r pois_cv_rmse` days (SD = `r pois_cv_sd`).
Given that `n_high_days` ranges from 0 to ~30, this reflects moderate
prediction error on the count scale.

**Limitation.** This RMSE is not directly comparable to any other metric
in this report: the unit of observation is a household–month and the
response is a raw count. Cross-model ranking would be misleading.

Across all six models, interpretability and predictive performance trade
off consistently. The LM, GAMs, and GLM Poisson offer transparent,
coefficient-based explanations but leave more variance unexplained. The
SVM and Neural Network improve accuracy but are effectively black boxes.
The GLM Binomial sits in between — log-odds coefficients remain
interpretable while classification performance is competitive. The right
choice depends on whether the use case demands explainability or accuracy.

# Conclusions {#conclusions}

Across 127 Swiss households monitored daily from 2022 to 2024, this analysis
identified the key drivers of heat pump electricity consumption and built a
foundation for demand modelling relevant to grid operators and energy suppliers.

Cold weather is the strongest structural driver of daily consumption, but its
effect is not uniform. Houses show a sharply accelerating response on the
coldest winter days, a pattern that has direct implications for load forecasting
during cold snaps and that a plain linear model understates. Beyond weather,
heat pump type and occupancy are the two most consistent predictors across all
models. Ground-source heat pumps consume approximately 13% less electricity per
day than air-source systems, and every additional resident adds roughly 14% to
daily demand regardless of building type or season.

For classifying unusually high-demand days, weekend occupancy is the single
strongest predictor, with cold weather second. Structural characteristics become
non-significant once the model accounts for each household's own baseline,
meaning demand spikes are driven by behaviour and weather rather than building
or system properties.

The GAM achieved the best predictive accuracy for daily consumption with a MAE
of 10.6 kWh per day and an R² of 31%, modestly improving on the Linear Model.
Both models leave roughly two thirds of day-to-day variation unexplained,
pointing to the value of incorporating behavioural signals in future work. For
classification, the Neural Network achieved the strongest performance with an
AUC of `r round(class_summary[grepl("Neural", class_summary$Model), "CV AUC"], 3)`
and accuracy of `r round(class_summary[grepl("Neural", class_summary$Model), "CV Accuracy"] * 100, 1)`%,
with the SVM close behind at an AUC of
`r round(class_summary[grepl("SVM", class_summary$Model), "CV AUC"], 3)`
and accuracy of
`r round(class_summary[grepl("SVM", class_summary$Model), "CV Accuracy"] * 100, 1)`%.
For a client prioritising transparency, the GLM Binomial is recommended,
achieving an AUC of
`r round(class_summary[grepl("GLM", class_summary$Model), "CV AUC"], 3)`
and accuracy of
`r round(class_summary[grepl("GLM", class_summary$Model), "CV Accuracy"] * 100, 1)`%
while remaining fully interpretable.

The practical implication for energy suppliers and grid operators is to target
demand management programmes at larger households with air-source systems during
cold weather peaks, where high occupancy and less efficient heating combine to
produce the sharpest load spikes. Weekend time-of-use tariffs or demand response
incentives offer an additional lever for flattening the load peaks that drive
balancing energy costs.

# Use of Generative AI {#ai}

Generative AI tools — primarily Claude (Anthropic) and GitHub Copilot — were used throughout this project to support code generation, debugging, and text drafting. AI assistance proved effective for producing boilerplate R code for data preprocessing, visualisation, and model-fitting wrappers, as well as for generating first drafts of methodological prose describing model outputs and statistical concepts. Human oversight was essential for tasks requiring analytical judgement, including the selection of modelling thresholds, cross-validation architecture, and the interpretation of model coefficients, where AI-generated suggestions occasionally contained subtle errors or overstated certainty. All AI-generated code was reviewed and tested prior to integration, and all interpretive text was verified against actual model output before inclusion in this report.
