---
output:
  pdf_document:
    toc: true
    toc_depth: '3'
  html_document:
    toc: true
    toc_float: true
    toc_depth: 3
    number_sections: true
    theme: flatly
    code_folding: hide
subtitle: Applied Machine Learning and Predictive Modelling 1 – FS 2026
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo    = TRUE,
  message = FALSE,
  warning = FALSE
)
library(ggplot2)
library(dplyr)
set.seed(1234)
```

## Poisson GLM {#poisson}
*Lead: Raphaël Spagolla*

```{r import-poisson, echo=FALSE}
dat_count  <- readRDS("../data_processed/poisson/dat_count.rds")
mod_pois   <- readRDS("../models/poisson_glm/mod_pois.rds")
mod_quasi  <- readRDS("../models/poisson_glm/mod_quasi.rds")
rr_table   <- read.csv("../models/poisson_glm/coeff_rate_ratios.csv")
dispersion <- readRDS("../models/poisson_glm/dispersion_val.rds")
comp       <- read.csv("../models/poisson_glm/quasi_vs_pois_se_pvals.csv")
```

### Objective

This section models the number of high-consumption days per household per month using a Poisson GLM. Unlike the binary approach used in the neural network section, this count-based formulation quantifies how frequently high-consumption events occur over a monthly horizon, which is more informative for resource planning, billing forecasting, and targeted energy-efficiency interventions.

The Poisson GLM is well-suited for this task: it is designed for non-negative integer outcomes and models the log of the expected count as a linear function of predictors. An offset of `log(n_days)` is included to account for varying observation windows across household-months, ensuring the model estimates rates rather than raw counts.


### Model Setup

```{r model-formula, echo=TRUE}
cat(deparse(formula(mod_pois)), "\n\n")
mod_pois$call
```

The Poisson GLM is specified with a log link and includes all available predictors: weather conditions (`heating_degree_days`, `temp_avg`), household characteristics (`living_area`, `building_type`, `heatpump_type`, `has_floor_heating`, `n_residents`), and a temporal usage indicator (`share_weekend`). The offset `log(n_days)` is included to model rates rather than raw counts.

The log link ensures that predicted counts are always non-negative. Each coefficient can be interpreted as the log of a rate ratio: the multiplicative change in the expected count rate associated with a one-unit increase in the predictor, holding all other variables constant.

### Model Results

```{r model-fit, echo=FALSE}
summ_p <- summary(mod_pois)

knitr::kable(
  data.frame(
    Metric = c("Null deviance", "Residual deviance", "AIC", "df residual"),
    Value  = c(
      round(summ_p$null.deviance, 2),
      round(summ_p$deviance, 2),
      round(summ_p$aic, 2),
      summ_p$df.residual
    )
  ),
  caption = "Poisson GLM fit statistics"
)
```

The model achieves a substantial reduction from null to residual deviance, indicating that the predictors collectively explain a meaningful share of variation in the count outcome. The AIC provides a basis for comparison with alternative specifications.

```{r coeff-table, echo=FALSE}
knitr::kable(
  rr_table,
  digits  = 4,
  caption = "Poisson GLM — Coefficients and Rate Ratios (exp(β)) with 95% Wald CI"
)
```

The rate ratio column (`rate_ratio`) provides the most interpretable summary of each predictor's effect. A rate ratio above 1 indicates a positive association with high-consumption frequency; a value below 1 indicates a negative association. Predictors whose 95% confidence interval (`rr_ci_low`, `rr_ci_high`) excludes 1 are statistically significant at the 5% level.

### Visual Evidence

```{r coeff-plot, echo=FALSE, fig.width=8, fig.height=6}
rr_plot <- rr_table %>%
  filter(term != "(Intercept)") %>%
  mutate(term = reorder(term, rate_ratio))

ggplot(rr_plot, aes(x = rate_ratio, y = term)) +
  geom_vline(xintercept = 1, linetype = "dashed", colour = "grey50") +
  geom_errorbarh(
    aes(xmin = rr_ci_low, xmax = rr_ci_high),
    height = 0.3, colour = "steelblue"
  ) +
  geom_point(size = 2.5, colour = "steelblue") +
  labs(
    title = "Rate ratios with 95% Wald CI (Poisson GLM)",
    x     = "Rate ratio (exp(β))",
    y     = NULL
  ) +
  theme_bw(base_size = 11)
```

The forest plot summarises the estimated rate ratios and their uncertainty. Predictors with confidence intervals entirely to the right of 1 are associated with higher high-consumption rates; those entirely to the left suggest a protective effect. The dashed line at 1 marks the null hypothesis of no association.

```{r residual-plots, echo=FALSE, fig.width=10, fig.height=4}
par(mfrow = c(1, 2))

# Pearson residuals vs fitted
plot(
  fitted(mod_pois),
  residuals(mod_pois, type = "pearson"),
  pch  = 16, cex = 0.3, col = rgb(0, 0, 0, 0.2),
  main = "Pearson residuals vs fitted values",
  xlab = "Fitted values",
  ylab = "Pearson residuals"
)
abline(h = 0, lty = 2, col = "tomato")

# Q-Q plot of Pearson residuals
qqnorm(
  residuals(mod_pois, type = "pearson"),
  pch  = 16, cex = 0.3, col = rgb(0, 0, 0, 0.2),
  main = "Q-Q plot of Pearson residuals"
)
qqline(residuals(mod_pois, type = "pearson"), col = "tomato", lty = 2)

par(mfrow = c(1, 1))
```

The residual plots provide a diagnostic check on model fit. In a well-specified Poisson GLM, Pearson residuals should be approximately centred around zero with no systematic pattern. Departure from the Q-Q line in the tails may indicate overdispersion or model misspecification.

### Model Diagnostics: Overdispersion

A key assumption of the Poisson GLM is that the conditional mean equals the conditional variance (equidispersion). In practice, count data often exhibit overdispersion — variance exceeding the mean — which inflates standard errors if left unaddressed.

```{r overdispersion, echo=FALSE}
knitr::kable(
  data.frame(
    Metric = c("Residual deviance", "Residual df", "Dispersion (deviance / df)"),
    Value  = c(
      round(summ_p$deviance, 2),
      summ_p$df.residual,
      round(dispersion, 4)
    )
  ),
  caption = "Overdispersion check"
)
```

The estimated dispersion ratio is `r round(dispersion, 3)`. `r if (dispersion > 1.5) " This value substantially exceeds 1, indicating overdispersion. A quasi-Poisson model is therefore fitted to obtain robust standard errors." else "This value is close to 1, suggesting that the equidispersion assumption is approximately satisfied."`

### Quasi-Poisson Comparison

```{r quasi-table, echo=FALSE}
knitr::kable(
  comp,
  digits  = 4,
  caption = "Poisson vs quasi-Poisson: standard errors and p-values"
)
```

```{r quasi-se-plot, echo=FALSE, fig.width=8, fig.height=5}
comp_long <- comp %>%
  filter(term != "(Intercept)") %>%
  mutate(term = reorder(term, se_quasi))

ggplot(comp_long, aes(y = term)) +
  geom_point(aes(x = se_pois,  colour = "Poisson"),       size = 2.5) +
  geom_point(aes(x = se_quasi, colour = "Quasi-Poisson"), size = 2.5) +
  scale_colour_manual(
    values = c("Poisson" = "steelblue", "Quasi-Poisson" = "tomato"),
    name   = "Model"
  ) +
  labs(
    title = "Standard errors: Poisson vs quasi-Poisson",
    x     = "Standard error",
    y     = NULL
  ) +
  theme_bw(base_size = 11)
```

The quasi-Poisson model uses the same point estimates as the standard Poisson model — the coefficients are identical — but inflates standard errors by a factor of $\sqrt{\hat{\phi}}$, where $\hat{\phi}$ is the estimated dispersion. This leads to wider confidence intervals and more conservative p-values, providing more reliable inference when overdispersion is present.

The estimated quasi-Poisson dispersion parameter is $\hat{\phi} =$ `r round(summary(mod_quasi)$dispersion, 3)`. `r if (summary(mod_quasi)$dispersion > 1.5) "The inflation of standard errors relative to the standard Poisson is visible in the comparison plot above, and p-values should be interpreted from the quasi-Poisson model." else "The adjustment is minor, indicating that the standard Poisson inference is largely reliable."`

### Key Findings

* The Poisson GLM successfully models monthly high-consumption day counts, with the offset ensuring that household-months with different numbers of observed days are treated on an equivalent rate basis.

* The model explains a substantial share of variation in the count outcome, as evidenced by the reduction from null to residual deviance.

* Rate ratios indicate which predictors are associated with higher or lower high-consumption rates. Weather variables, in particular temperature, show the expected direction of effect.

* The dispersion ratio of `r round(dispersion, 3)` `r if (dispersion > 1.5) "indicates overdispersion, making the quasi-Poisson model the preferred basis for inference." else "is close to 1, supporting the standard Poisson as a valid model."` 

* The quasi-Poisson model yields identical point estimates but corrected standard errors, ensuring reliable hypothesis tests even under moderate overdispersion.

### Conclusion

The Poisson GLM provides a principled and interpretable framework for modelling the frequency of high-consumption events at the household-month level. By using an offset to account for exposure, the model estimates rates rather than raw counts, enabling fair comparison across months with differing observation windows.

The estimated rate ratios offer clear, interpretable insights into which household and environmental characteristics are associated with higher high-consumption rates. These findings can directly inform targeted interventions such as energy-efficiency programmes or demand forecasting.

The overdispersion check reveals that `r if (dispersion > 1.5) "the variance exceeds the mean — a common feature of real-world count data — and the quasi-Poisson adjustment ensures that uncertainty is not underestimated." else "the equidispersion assumption is approximately satisfied, supporting the validity of the standard Poisson framework."` The quasi-Poisson results are therefore reported alongside the standard model as the basis for inference.

Further model improvements could include zero-inflated Poisson or negative binomial specifications if excess zeros or severe overdispersion remain a concern, as well as the inclusion of household-level random effects to account for repeated measurements across months.
