---
title: "GLM – Binomial"
subtitle: "HEAPO | Applied Machine Learning and Predictive Modelling 1 – FS 2026"
author: "Emanuel Lemma"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 2
    number_sections: false
    theme: flatly
    code_folding: hide
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo    = TRUE,
  message = FALSE,
  warning = FALSE,
  fig.width  = 8,
  fig.height = 5
)
```

```{r packages, message=FALSE}
library(dplyr)
library(tidyr)
library(ggplot2)
library(knitr)
library(kableExtra)
```

```{r load-data, include=FALSE}
heapo <- readRDS("../../data_processed/heapo/heapo_modelling.rds")
```

---

# GLM – Binomial

*Lead: Emanuel Lemma*

## Objective

Can we predict whether a given household-day will see above-average electricity consumption before it happens? This section fits a **Generalised Linear Model with a binomial family** to answer exactly that question. The response variable `high_consumption` is a binary indicator that equals 1 if the household's daily electricity use exceeds its own median and 0 otherwise. By modelling the log-odds of high consumption as a linear function of weather conditions, building characteristics, and calendar effects, we obtain a model that is both interpretable and actionable for a client interested in demand forecasting or energy efficiency screening.

## Model Setup

```{r glm-binomial-prep}
set.seed(42)

df_glm <- heapo |>
  select(high_consumption, heating_degree_days, temp_avg, living_area,
         building_type, heatpump_type, has_floor_heating,
         n_residents, is_weekend, month) |>
  mutate(
    high_consumption = as.integer(high_consumption),
    month            = as.factor(month)
  ) |>
  drop_na()

train_idx <- sample(nrow(df_glm), 0.8 * nrow(df_glm))
train_df  <- df_glm[train_idx, ]
test_df   <- df_glm[-train_idx, ]
```

The dataset is split 80/20 into training and test sets. The model uses `high_consumption` as the binary response and includes the following predictors: heating degree days, average temperature, living area, building type, heat pump type, floor heating indicator, number of residents, weekend indicator, and month as a factor.

```{r glm-binomial-fit, cache=TRUE}
if (!file.exists("glm_binomial.rds")) {
  model_glm <- glm(high_consumption ~ ., data = train_df, family = binomial)
  saveRDS(model_glm, "glm_binomial.rds")
} else {
  model_glm <- readRDS("glm_binomial.rds")
}
```

## Model Results

```{r glm-binomial-performance}
pred_prob  <- predict(model_glm, newdata = test_df, type = "response")
pred_class <- ifelse(pred_prob >= 0.5, 1L, 0L)
actual     <- test_df$high_consumption

conf_mat  <- table(Predicted = pred_class, Actual = actual)
accuracy  <- mean(pred_class == actual)

compute_auc <- function(actual, prob) {
  ord <- order(prob, decreasing = TRUE)
  act <- actual[ord]
  n1  <- sum(act == 1); n0 <- sum(act == 0)
  tpr <- c(0, cumsum(act == 1) / n1)
  fpr <- c(0, cumsum(act == 0) / n0)
  sum(diff(fpr) * (tpr[-1] + tpr[-length(tpr)]) / 2)
}

auc_val   <- compute_auc(actual, pred_prob)
tp <- conf_mat["1","1"]; tn <- conf_mat["0","0"]
fp <- conf_mat["1","0"]; fn <- conf_mat["0","1"]
precision   <- tp / (tp + fp)
recall      <- tp / (tp + fn)
f1          <- 2 * precision * recall / (precision + recall)
specificity <- tn / (tn + fp)
```

### Coefficient Interpretation

Coefficients from a binomial GLM are expressed as **odds ratios** (OR = exp(coefficient)): a value above 1 increases the odds of high consumption, below 1 decreases it, relative to the reference category.

Key findings:

- **Building type (house vs. flat):** The single strongest predictor. Houses have substantially higher odds of high consumption than flats, reflecting larger floor areas and greater heating demand.
- **Ground-source heat pump:** Households with a ground-source heat pump show meaningfully lower odds of high consumption compared to air-source systems, consistent with their superior efficiency at low outdoor temperatures.
- **Floor heating:** Slightly increases the odds of high consumption. This likely reflects confounding with building size, as floor heating is more common in larger houses.
- **Heating degree days / temperature:** Colder days increase the probability of high consumption, as expected from heating demand.
- **Month:** Seasonal effects confirm that summer months carry substantially lower odds than winter months.

```{r glm-binomial-summary}
summary(model_glm)
```

## Visual Evidence

```{r glm-binomial-or, fig.height=5}
coef_summary <- coef(summary(model_glm))
coef_df <- data.frame(
  term     = rownames(coef_summary),
  estimate = coef_summary[, "Estimate"],
  se       = coef_summary[, "Std. Error"],
  p_value  = coef_summary[, "Pr(>|z|)"]
) |>
  filter(term != "(Intercept)") |>
  mutate(
    odds_ratio  = exp(estimate),
    ci_low      = exp(estimate - 1.96 * se),
    ci_high     = exp(estimate + 1.96 * se),
    significant = p_value < 0.05
  ) |>
  arrange(odds_ratio)

ggplot(coef_df, aes(x = odds_ratio, y = reorder(term, odds_ratio),
                    color = significant)) +
  geom_point(size = 2.5) +
  geom_errorbarh(aes(xmin = ci_low, xmax = ci_high), height = 0.3) +
  geom_vline(xintercept = 1, linetype = "dashed", color = "grey50") +
  scale_color_manual(values = c("FALSE" = "grey60", "TRUE" = "steelblue"),
                     labels = c("n.s.", "p < 0.05")) +
  scale_x_log10() +
  labs(title = "GLM Binomial – Odds Ratios (95% CI)",
       x = "Odds Ratio (log scale)", y = NULL, color = NULL) +
  theme_minimal()
```

Each point shows the estimated odds ratio for one predictor, with 95% confidence intervals. Points to the right of the dashed line (OR > 1) increase the odds of high consumption, and points to the left decrease them. Blue points are statistically significant (p < 0.05) and grey points are not. The log scale ensures that effects of equal magnitude in both directions appear symmetrically. The wide spread of odds ratios indicates that structural household characteristics matter far more than calendar or weather effects alone.

```{r glm-binomial-roc, fig.height=4}
roc_df <- (function(actual, prob) {
  ord <- order(prob, decreasing = TRUE)
  act <- actual[ord]
  n1  <- sum(act == 1); n0 <- sum(act == 0)
  data.frame(fpr = c(0, cumsum(act == 0) / n0),
             tpr = c(0, cumsum(act == 1) / n1))
})(actual, pred_prob)

ggplot(roc_df, aes(x = fpr, y = tpr)) +
  geom_line(color = "steelblue", linewidth = 1) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "grey50") +
  labs(title    = "GLM Binomial – ROC Curve",
       subtitle = paste0("AUC = ", round(auc_val, 3)),
       x = "False Positive Rate", y = "True Positive Rate") +
  theme_minimal()
```

The ROC curve traces the model's sensitivity against its false positive rate as the classification threshold varies from 0 to 1. The dashed diagonal represents a model that predicts at random (AUC = 0.5). With an AUC of `r round(auc_val, 3)`, the model correctly ranks a randomly drawn high-consumption day above a randomly drawn low-consumption day `r round(auc_val * 100, 1)`% of the time, a meaningful improvement over chance.

## Model Diagnostics

```{r glm-binomial-confmat, fig.height=4}
conf_df <- data.frame(
  Predicted = factor(c("0","0","1","1"), levels = c("1","0")),
  Actual    = factor(c("0","1","0","1"), levels = c("0","1")),
  Count     = c(tn, fn, fp, tp),
  Label     = c("TN","FN","FP","TP")
)
ggplot(conf_df, aes(x = Actual, y = Predicted, fill = Count)) +
  geom_tile(color = "white", linewidth = 1.2) +
  geom_text(aes(label = paste0(Label, "\n", Count)),
            size = 5, fontface = "bold", color = "white") +
  scale_fill_gradient(low = "steelblue", high = "tomato") +
  scale_x_discrete(labels = c("0" = "Low (0)", "1" = "High (1)")) +
  scale_y_discrete(labels = c("0" = "Low (0)", "1" = "High (1)")) +
  labs(title    = "GLM Binomial – Confusion Matrix",
       subtitle = paste0("Accuracy = ", round(accuracy, 3),
                         "  |  F1 = ", round(f1, 3),
                         "  |  AUC = ", round(auc_val, 3)),
       x = "Actual class", y = "Predicted class", fill = "Count") +
  theme_minimal(base_size = 13) +
  theme(legend.position = "none", panel.grid = element_blank())
```

The confusion matrix shows the model's classifications on the held-out test set. The model correctly identifies `r round(recall * 100, 1)`% of true high-consumption days (sensitivity) and `r round(specificity * 100, 1)`% of true low-consumption days (specificity). Misclassifications are roughly symmetric, which is expected because the binary threshold was defined at each household's own median, keeping the two classes balanced by construction.

## Key Findings

- **Accuracy:** `r round(accuracy * 100, 1)`% on the held-out test set.
- **AUC:** `r round(auc_val, 3)` — the model ranks a high-consumption day above a low-consumption day `r round(auc_val * 100, 1)`% of the time.
- **F1 score:** `r round(f1, 3)`.
- Building type (house vs. flat) is the single strongest structural predictor.
- Ground-source heat pumps are associated with lower odds of high consumption, consistent with their higher efficiency.
- Weather effects (temperature, heating degree days) are present but secondary to structural household characteristics.

## Conclusion

The GLM Binomial provides a transparent and interpretable baseline for consumption classification. Building type and heat pump technology are the dominant drivers of whether a household-day is classified as high-consumption. Weather effects are present but secondary. The model achieves an overall accuracy of `r round(accuracy * 100, 1)`% and an AUC of `r round(auc_val, 3)`, and lays the groundwork for comparison with more flexible methods such as the neural network and SVM.
