---
title: "HEAPO – Data Preparation (Daily Resolution)"
author: ""
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    code_folding: show
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
knitr::opts_knit$set(root.dir = normalizePath(".."))

library(dplyr)
library(readr)
library(tidyr)
library(lubridate)
library(purrr)
library(ggplot2)

# Paths relative to project root
base_path   <- "data_raw/heapo_data"
output_path <- "data_processed/heapo"
```

## Objective

This document describes the preparation of the HEAPO dataset at **daily resolution**. The resulting dataset contains one row per household and day, with daily electricity consumption and associated weather and household variables.

---

## 1. Load Metadata

```{r meta}
meta <- read_delim(file.path(base_path, "meta_data", "meta_data.csv"),
                   delim = ";", show_col_types = FALSE) %>%
  mutate(Household_ID = as.character(Household_ID))

households <- read_delim(file.path(base_path, "meta_data", "households.csv"),
                         delim = ";", show_col_types = FALSE) %>%
  mutate(Household_ID = as.character(Household_ID))

household_meta <- meta %>%
  left_join(households, by = "Household_ID") %>%
  transmute(
    Household_ID,
    weather_region       = Weather_ID,
    group                = factor(Group),
    building_type        = factor(Survey_Building_Type),
    heatpump_type        = factor(Survey_HeatPump_Installation_Type),
    living_area          = as.numeric(Survey_Building_LivingArea),
    n_residents          = as.integer(Survey_Building_Residents),
    has_floor_heating    = as.integer(as.logical(Survey_HeatDistribution_System_FloorHeating)),
    has_radiator         = as.integer(as.logical(Survey_HeatDistribution_System_Radiator)),
    has_dryer            = as.integer(as.logical(Survey_Installation_HasDryer)),
    has_freezer          = as.integer(as.logical(Survey_Installation_HasFreezer)),
    has_electric_vehicle = as.integer(as.logical(Survey_Installation_HasElectricVehicle))
  ) %>%
  mutate(
    has_electric_vehicle = ifelse(is.na(has_electric_vehicle), 0L, has_electric_vehicle),
    has_dryer            = ifelse(is.na(has_dryer),            0L, has_dryer),
    has_freezer          = ifelse(is.na(has_freezer),          0L, has_freezer)
  )

cat("Households with metadata:", nrow(household_meta))
```

---

## 2. Load Daily Smart Meter Data

```{r smartmeter, cache=TRUE}
daily_dir   <- file.path(base_path, "smart_meter_data", "daily")
daily_files <- list.files(daily_dir, pattern = "\\.csv$", full.names = TRUE)

cat("Number of files:", length(daily_files))

daily_consumption <- map_dfr(daily_files, function(f) {
  df <- read_delim(f, delim = ";", show_col_types = FALSE,
                   col_types = cols(.default = col_character()))
  if (nrow(df) == 0) return(NULL)

  df %>%
    transmute(
      Household_ID = as.character(Household_ID),
      date         = as.Date(substr(Timestamp, 1, 10)),
      kWh_total    = as.numeric(kWh_received_Total),
      kWh_heatpump = as.numeric(kWh_received_HeatPump),
      kWh_other    = as.numeric(kWh_received_Other),
      kWh_returned = as.numeric(kWh_returned_Total)
    ) %>%
    filter(!is.na(date), !is.na(kWh_total), kWh_total >= 0)
})

cat("Entries:", nrow(daily_consumption))
cat("Households:", n_distinct(daily_consumption$Household_ID))
cat("Time period:", as.character(min(daily_consumption$date)),
    "to", as.character(max(daily_consumption$date)))
```

---

## 3. Load Daily Weather Data

```{r weather}
weather_daily_dir <- file.path(base_path, "weather_data", "daily")
weather_files     <- list.files(weather_daily_dir, pattern = "\\.csv$", full.names = TRUE)

weather_daily <- map_dfr(weather_files, function(f) {
  read_delim(f, delim = ";", show_col_types = FALSE) %>%
    transmute(
      weather_region      = Weather_ID,
      date                = as.Date(Timestamp),
      temp_max            = Temperature_max_daily,
      temp_min            = Temperature_min_daily,
      temp_avg            = Temperature_avg_daily,
      temp_range          = Temperature_max_daily - Temperature_min_daily,
      heating_degree_days = HeatingDegree_SIA_daily,
      cooling_degree_days = CoolingDegree_US_daily,
      humidity_avg        = Humidity_avg_daily,
      precipitation       = Precipitation_total_daily,
      sunshine_hours      = Sunshine_duration_daily
    )
})

cat("Weather entries:", nrow(weather_daily))
```

---

## 4. Merge

```{r join}
df <- daily_consumption %>%
  inner_join(household_meta, by = "Household_ID") %>%
  left_join(weather_daily, by = c("weather_region", "date"))

cat("Rows after join:", nrow(df))
```

---

## 5. Feature Engineering

```{r features}
df <- df %>%
  mutate(
    year       = year(date),
    month      = month(date),
    weekday    = wday(date, label = FALSE, week_start = 1),
    is_weekend = as.integer(weekday >= 6),
    kWh_per_m2 = ifelse(!is.na(living_area) & living_area > 0,
                        kWh_total / living_area, NA)
  )
```

---

## 6. Cleaning & Final Dataset

```{r clean}
core_vars <- c("kWh_total", "living_area", "n_residents",
               "building_type", "heatpump_type", "weather_region",
               "has_floor_heating", "has_radiator", "temp_avg")

heapo_daily <- df %>%
  drop_na(all_of(core_vars)) %>%
  select(
    Household_ID, date, year, month, weekday, is_weekend,
    kWh_total, kWh_heatpump, kWh_other, kWh_returned, kWh_per_m2,
    temp_avg, temp_min, temp_max, temp_range,
    heating_degree_days, cooling_degree_days,
    humidity_avg, precipitation, sunshine_hours,
    weather_region, group, building_type, heatpump_type,
    living_area, n_residents,
    has_floor_heating, has_radiator, has_dryer, has_freezer, has_electric_vehicle
  )

cat("Final dataset:", nrow(heapo_daily), "rows,", ncol(heapo_daily), "variables")
cat("Households:", n_distinct(heapo_daily$Household_ID))
cat("Time period:", as.character(min(heapo_daily$date)),
    "to", as.character(max(heapo_daily$date)))
```

---

## 7. Dataset Overview

### Variable Types

```{r overview}
glimpse(heapo_daily)
```

### Missing Values

```{r missing}
na_pct <- sapply(heapo_daily, function(x) round(mean(is.na(x)) * 100, 1))
na_pct[na_pct > 0]
```

### Consumption over Time

```{r plot-time, fig.width=10, fig.height=4}
heapo_daily %>%
  group_by(date) %>%
  summarise(kWh_median = median(kWh_total)) %>%
  ggplot(aes(x = date, y = kWh_median)) +
  geom_line(alpha = 0.6, color = "steelblue") +
  labs(title = "Median Daily Consumption Across All Households",
       x = "Date", y = "kWh (Median)") +
  theme_minimal()
```

### Consumption by Heat Pump Type

```{r plot-type, fig.width=7, fig.height=4}
ggplot(heapo_daily, aes(x = heatpump_type, y = kWh_total, fill = heatpump_type)) +
  geom_boxplot(outlier.alpha = 0.2) +
  labs(title = "Daily Consumption by Heat Pump Type",
       x = "", y = "kWh") +
  theme_minimal() +
  theme(legend.position = "none")
```

---

## 8. Save

```{r save}
saveRDS(heapo_daily, file.path(output_path, "heapo_dataset_daily.rds"))
write_csv(heapo_daily, file.path(output_path, "heapo_dataset_daily.csv"))

cat("Saved as heapo_dataset_daily.rds and .csv")
```
