---
title: "HEAPO – Data Filtration & Exploratory Data Analysis"
author: ""
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float:
      collapsed: false
    code_folding: hide
    theme: flatly
    df_print: paged
    number_sections: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = FALSE,
  message = FALSE,
  warning = FALSE,
  fig.width = 10,
  fig.height = 5
)
knitr::opts_knit$set(root.dir = normalizePath(".."))

library(dplyr)
library(tidyr)
library(ggplot2)
library(lubridate)
library(knitr)
library(kableExtra)
library(scales)

set.seed(42)

styled_table <- function(df, caption = "") {
  df %>%
    kable(caption = caption) %>%
    kable_styling(
      bootstrap_options = c("striped", "hover", "condensed"),
      full_width = FALSE
    )
}
```

```{r load-data, include=FALSE}
heapo_daily <- readRDS("data_processed/heapo/heapo_dataset_daily.rds")
```

---

# Data Filtration

The full daily dataset spans 2019–2024 and contains data from up to 756 households, totalling over 900,000 observations. This is too large for methods such as Neural Networks and SVMs (practical limit: ~100,000 rows). The filtration procedure defined below was derived from the findings in the **Data Quality Report** (`03_data_quality_report.Rmd`).

## Step 1: Filter Time Window

The time window **2022-01-01 to 2024-02-29** was chosen because:

- Only 17 households have complete coverage over the full period (2019–2024), making a full-period subset infeasible.
- From 2022 onwards, 403 households have ≥95% coverage, forming a large and representative pool.
- The window covers two full calendar years plus two months, capturing all seasonal cycles at least twice (critical for heating-related modelling).

```{r filter-time}
window_start <- as.Date("2022-01-01")
window_end <- as.Date("2024-02-29")

heapo_window <- heapo_daily %>%
  filter(date >= window_start, date <= window_end)

days_in_window <- as.numeric(window_end - window_start) + 1

cat("Rows after time filter:", nrow(heapo_window), "\n")
cat("Households in window:  ", n_distinct(heapo_window$Household_ID), "\n")
cat("Days in window:        ", days_in_window, "\n")
```

## Step 2: Identify Eligible Households (≥95% Coverage)

A household is considered *eligible* if it reported data on at least 95% of the days in the time window. This threshold ensures complete seasonal coverage while tolerating minor gaps (e.g., a few missing days due to meter downtime).

```{r filter-coverage}
coverage <- heapo_window %>%
  group_by(Household_ID) %>%
  summarise(n_days = n(), .groups = "drop") %>%
  mutate(coverage_pct = n_days / days_in_window * 100)

eligible <- coverage %>% filter(coverage_pct >= 95)

cat("Eligible households (>=95% coverage):", nrow(eligible), "\n")
```

> **Finding:** `r nrow(eligible)` households meet the ≥95% coverage threshold in the 2022–2024 window, consistent with the quality report's finding of 403 eligible households.

## Step 3: Random Sample of Households (Max. 100,000 Rows)

To stay within the 100,000 observation limit, eligible households are randomly shuffled and added cumulatively until the row limit is reached. Each eligible household contributes approximately 788 days, so roughly 126 households fit within the limit.

```{r filter-sample}
# Randomly shuffle eligible households, then take as many as fit under 100,000 rows
sampled_ids <- eligible %>%
  slice_sample(n = nrow(eligible)) %>% # random shuffle
  mutate(cumulative_rows = cumsum(n_days)) %>%
  filter(cumulative_rows <= 100000) %>%
  pull(Household_ID)

heapo_model <- heapo_window %>%
  filter(Household_ID %in% sampled_ids)

cat("Selected households:", n_distinct(heapo_model$Household_ID), "\n")
cat("Observations:       ", nrow(heapo_model), "\n")
cat(
  "Date range:         ",
  as.character(min(heapo_model$date)),
  "to",
  as.character(max(heapo_model$date)),
  "\n"
)
stopifnot(nrow(heapo_model) <= 100000) # hard check
```

## Step 4: Derive Response Variables

Two derived response variables are added for use in GLM Binomial and other methods:

- **`high_consumption`**: 1 if `kWh_total` exceeds the dataset-wide median, 0 otherwise (used as binary response in GLM Binomial).
- **`log_kWh_total`**: natural log of `kWh_total` (mandatory for LM, as `kWh_total` is an amount variable with right-skewed distribution).

```{r derive-vars}
median_kWh <- median(heapo_model$kWh_total, na.rm = TRUE)

heapo_model <- heapo_model %>%
  mutate(
    high_consumption = as.integer(kWh_total > median_kWh),
    log_kWh_total = log(kWh_total)
  )

cat("Median kWh_total used for threshold:", round(median_kWh, 3), "\n")
cat(
  "high_consumption = 1:",
  sum(heapo_model$high_consumption),
  "(",
  round(mean(heapo_model$high_consumption) * 100, 1),
  "%)\n"
)
```

## Step 5: Save Final Modelling Dataset

```{r save}
saveRDS(heapo_model, "data_processed/heapo/heapo_modelling.rds")
write.csv(
  heapo_model,
  "data_processed/heapo/heapo_modelling.csv",
  row.names = FALSE
)
cat("Saved: data_processed/heapo/heapo_modelling.rds/.csv\n")
```

## Filtration Summary

```{r filtration-summary}
tibble(
  Step = c(
    "0 – Full daily dataset",
    "1 – After time window filter (2022-01-01 to 2024-02-29)",
    "2 – Eligible households (>=95% coverage)",
    paste0("3 – Random sample (n = ", length(sampled_ids), ", max 100k rows)"),
    "Final modelling dataset"
  ),
  Households = c(
    n_distinct(heapo_daily$Household_ID),
    n_distinct(heapo_window$Household_ID),
    nrow(eligible),
    length(sampled_ids),
    n_distinct(heapo_model$Household_ID)
  ),
  Observations = c(
    nrow(heapo_daily),
    nrow(heapo_window),
    NA,
    NA,
    nrow(heapo_model)
  )
) %>%
  styled_table("Filtration steps overview")
```

---

# Exploratory Data Analysis

The EDA is performed on the **final modelling dataset** (`heapo_model`: `r n_distinct(heapo_model$Household_ID)` households, `r comma(nrow(heapo_model))` observations, 2022–2024).

---

## Response Variable: kWh_total

### Distribution

```{r eda-response-dist-raw, fig.height=4}
ggplot(heapo_model, aes(x = kWh_total)) +
  geom_histogram(bins = 60, fill = "steelblue", color = "white") +
  scale_x_continuous(labels = comma) +
  labs(
    title = "Distribution of kWh_total (raw)",
    x = "kWh per day",
    y = "Count"
  ) +
  theme_minimal()
```

```{r eda-response-dist-log, fig.height=4}
ggplot(heapo_model, aes(x = log_kWh_total)) +
  geom_histogram(bins = 60, fill = "darkorange", color = "white") +
  labs(
    title = "Distribution of log(kWh_total)",
    x = "log(kWh per day)",
    y = "Count"
  ) +
  theme_minimal()
```

> **Interpretation:** Raw `kWh_total` is strongly right-skewed: a small number of days show very high consumption (cold winter peaks). After log-transformation, the distribution is approximately normal. This confirms that **`log(kWh_total)` must be used as the response in Linear Models**.

### Summary Statistics

```{r eda-response-stats}
heapo_model %>%
  summarise(
    Min = round(min(kWh_total), 2),
    Q1 = round(quantile(kWh_total, 0.25), 2),
    Median = round(median(kWh_total), 2),
    Mean = round(mean(kWh_total), 2),
    Q3 = round(quantile(kWh_total, 0.75), 2),
    Max = round(max(kWh_total), 2),
    SD = round(sd(kWh_total), 2)
  ) %>%
  styled_table("Summary statistics: kWh_total")
```

---

## Temporal Patterns

### Monthly Consumption

```{r eda-monthly, fig.height=5}
heapo_model %>%
  mutate(month_label = month(date, label = TRUE, abbr = TRUE)) %>%
  ggplot(aes(x = month_label, y = kWh_total)) +
  geom_boxplot(fill = "steelblue", alpha = 0.7, outlier.size = 0.3) +
  scale_y_continuous(labels = comma) +
  labs(
    title = "Daily kWh Consumption by Month",
    subtitle = "Strong seasonal pattern: winter months dominate",
    x = "Month",
    y = "kWh per day"
  ) +
  theme_minimal()
```

> **Interpretation:** There is a clear seasonal pattern driven by heating demand. Winter months (December, January, February) show median daily consumption roughly 3 to 5 times higher than summer months (June to August). This is expected: heat pumps are the dominant electricity consumer and operate mostly in winter. The wide interquartile range in winter reflects household heterogeneity in building insulation, size, and heat pump type.

### Weekday vs. Weekend

```{r eda-weekend, fig.height=4}
heapo_model %>%
  mutate(Day_Type = ifelse(is_weekend == 1, "Weekend", "Weekday")) %>%
  ggplot(aes(x = Day_Type, y = kWh_total, fill = Day_Type)) +
  geom_boxplot(alpha = 0.7, show.legend = FALSE) +
  scale_fill_manual(values = c("Weekday" = "steelblue", "Weekend" = "tomato")) +
  scale_y_continuous(labels = comma) +
  labs(
    title = "Daily Consumption: Weekday vs. Weekend",
    x = NULL,
    y = "kWh per day"
  ) +
  theme_minimal()
```

> **Interpretation:** Weekend consumption is slightly higher than weekday consumption on average, consistent with occupants being home more during weekends. However, the difference is modest relative to the seasonal signal, as heating demand is the primary driver of variation.

### Annual Trend

```{r eda-trend, fig.height=4}
heapo_model %>%
  group_by(date) %>%
  summarise(median_kWh = median(kWh_total), .groups = "drop") %>%
  ggplot(aes(x = date, y = median_kWh)) +
  geom_line(color = "steelblue", linewidth = 0.4, alpha = 0.8) +
  geom_smooth(
    method = "loess",
    span = 0.1,
    color = "tomato",
    linewidth = 1,
    se = FALSE
  ) +
  scale_y_continuous(labels = comma) +
  labs(
    title = "Median Daily kWh Consumption Over Time",
    subtitle = "Red line = LOESS smooth",
    x = "Date",
    y = "Median kWh per day"
  ) +
  theme_minimal()
```

> **Interpretation:** The time series confirms the seasonal cycle across two full years. Consumption peaks in January/February and drops sharply in summer. The amplitude and shape of the two visible cycles (2022 and 2023) are broadly similar, suggesting no major structural break in the data over this period.

---

## Weather Relationships

### Temperature vs. Consumption

```{r eda-temp, fig.height=5}
heapo_model %>%
  sample_n(min(20000, nrow(.))) %>% # subsample for plotting speed
  ggplot(aes(x = temp_avg, y = kWh_total)) +
  geom_point(alpha = 0.08, size = 0.5, color = "steelblue") +
  geom_smooth(method = "loess", color = "tomato", se = FALSE, linewidth = 1) +
  scale_y_continuous(labels = comma) +
  labs(
    title = "Average Temperature vs. Daily kWh Consumption",
    subtitle = "Each point = one household-day (random subsample of 20k)",
    x = "Temperature (°C)",
    y = "kWh per day"
  ) +
  theme_minimal()
```

> **Interpretation:** There is a strong non-linear negative relationship between temperature and consumption. Consumption rises sharply below roughly 15°C (the typical heating threshold) and levels off above it. This non-linearity justifies using GAM (which can capture this curve) over a simple linear model for temperature. Note the wide scatter at low temperatures, reflecting household-level heterogeneity.

### Heating Degree Days vs. Consumption

```{r eda-hdd, fig.height=5}
heapo_model %>%
  sample_n(min(20000, nrow(.))) %>%
  ggplot(aes(x = heating_degree_days, y = kWh_total)) +
  geom_point(alpha = 0.08, size = 0.5, color = "steelblue") +
  geom_smooth(method = "lm", color = "tomato", se = FALSE, linewidth = 1) +
  scale_y_continuous(labels = comma) +
  labs(
    title = "Heating Degree Days vs. Daily kWh Consumption",
    subtitle = "Heating Degree Days (HDD): 0 on days above the base temperature",
    x = "Heating Degree Days",
    y = "kWh per day"
  ) +
  theme_minimal()
```

> **Interpretation:** Heating Degree Days (HDD) measure the cumulative temperature deficit below a defined base temperature (here: 12°C per the Swiss SIA norm), giving a daily index of how much heating is required. A day with an average temperature of 2°C contributes 10 HDD; a day above 12°C contributes 0. HDD shows a strong positive linear relationship with consumption and is generally a stronger predictor than raw temperature, because it directly quantifies heating demand rather than ambient conditions.

---

## Household-Level Variation

### Consumption by Building Type

```{r eda-building, fig.height=5}
heapo_model %>%
  ggplot(aes(x = building_type, y = kWh_total, fill = building_type)) +
  geom_boxplot(alpha = 0.7, show.legend = FALSE, outlier.size = 0.3) +
  scale_y_continuous(labels = comma) +
  labs(
    title = "Daily Consumption by Building Type",
    x = "Building Type",
    y = "kWh per day"
  ) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 25, hjust = 1))
```

> **Interpretation:** Building type is associated with notable differences in median consumption. Larger or older building types (e.g., detached houses) tend to show higher and more variable consumption, reflecting differences in insulated floor area and construction era. This confirms `building_type` as a relevant predictor.

### Consumption by Heat Pump Type

```{r eda-hptype, fig.height=4}
heapo_model %>%
  ggplot(aes(x = heatpump_type, y = kWh_total, fill = heatpump_type)) +
  geom_boxplot(alpha = 0.7, show.legend = FALSE, outlier.size = 0.3) +
  scale_y_continuous(labels = comma) +
  labs(
    title = "Daily Consumption by Heat Pump Type",
    x = "Heat Pump Type",
    y = "kWh per day"
  ) +
  theme_minimal()
```

> **Interpretation:** The HEAPO dataset distinguishes two main heat pump types: **air-source** (extracts heat from outdoor air) and **ground/water-source** (extracts heat from the ground or groundwater). Air-source heat pumps are less efficient at low outdoor temperatures because the temperature difference between the heat source and the heating circuit increases, requiring more electrical work. This efficiency is measured by the **Coefficient of Performance (COP)**: the ratio of heat delivered to electricity consumed. A COP of 3 means 3 kWh of heat per 1 kWh of electricity. Ground-source heat pumps maintain a more stable COP year-round because ground temperatures remain relatively constant. Any consumption difference visible here may be partially confounded by building size, insulation, and climate zone.

### Living Area vs. Consumption

```{r eda-area, fig.height=5}
heapo_model %>%
  distinct(Household_ID, .keep_all = TRUE) %>%
  left_join(
    heapo_model %>%
      group_by(Household_ID) %>%
      summarise(median_kWh = median(kWh_total), .groups = "drop"),
    by = "Household_ID"
  ) %>%
  ggplot(aes(x = living_area, y = median_kWh)) +
  geom_point(alpha = 0.6, color = "steelblue") +
  geom_smooth(method = "lm", color = "tomato", se = TRUE, linewidth = 1) +
  scale_y_continuous(labels = comma) +
  labs(
    title = "Living Area vs. Median Daily Consumption (per household)",
    subtitle = "Each point = one household (median kWh across all its days)",
    x = "Living Area (m²)",
    y = "Median kWh per day"
  ) +
  theme_minimal()
```

> **Interpretation:** Larger living areas are associated with higher median consumption at the household level. The relationship is approximately linear, supporting inclusion of `living_area` as a continuous predictor. There is notable scatter: households with similar floor area can differ substantially in consumption, likely due to insulation quality, heat pump type, and occupant behaviour.

---

## Categorical Predictors

### Weather Region Distribution

```{r eda-region, fig.height=4}
heapo_model %>%
  distinct(Household_ID, weather_region) %>%
  count(weather_region) %>%
  ggplot(aes(x = reorder(weather_region, n), y = n, fill = n)) +
  geom_col(show.legend = FALSE) +
  geom_text(aes(label = n), hjust = -0.2) +
  coord_flip() +
  scale_fill_gradient(low = "lightblue", high = "steelblue") +
  labs(
    title = "Households per Weather Region (modelling subset)",
    x = "Weather Region",
    y = "Number of households"
  ) +
  theme_minimal()
```

> **Interpretation:** The sampled households are distributed across multiple weather regions. The distribution reflects the original dataset's geographic composition in the canton of Zurich, Switzerland. Some regions have fewer households, so the categorical predictor `weather_region` may need grouping or regularisation in certain models (e.g., SVM).

### Binary Predictors Overview

```{r eda-binary, fig.height=4}
binary_vars <- c(
  "has_floor_heating",
  "has_radiator",
  "has_dryer",
  "has_freezer",
  "has_electric_vehicle"
)

heapo_model %>%
  distinct(Household_ID, .keep_all = TRUE) %>%
  summarise(across(all_of(binary_vars), ~ round(mean(.) * 100, 1))) %>%
  pivot_longer(everything(), names_to = "Variable", values_to = "Pct_Yes") %>%
  ggplot(aes(x = reorder(Variable, Pct_Yes), y = Pct_Yes, fill = Pct_Yes)) +
  geom_col(show.legend = FALSE) +
  geom_text(aes(label = paste0(Pct_Yes, "%")), hjust = -0.2) +
  coord_flip() +
  scale_fill_gradient(low = "lightblue", high = "steelblue") +
  scale_y_continuous(limits = c(0, 100)) +
  labs(
    title = "Binary Predictors: % of Households with Value = 1",
    x = NULL,
    y = "% of households"
  ) +
  theme_minimal()
```

> **Interpretation:** Floor heating (radiant underfloor heating) distributes heat via pipes embedded in the floor and operates at lower supply temperatures (typically 30-40°C) compared to radiators (60-80°C). Heat pumps are more efficient at lower supply temperatures because the temperature lift (the difference between source and output) is smaller, resulting in a higher COP. Households with floor heating therefore tend to use electricity more efficiently for the same amount of heat delivered. Variables with very low prevalence (e.g., below 5%) may have limited predictive power or require careful handling in models.

---

## Correlation Analysis

```{r eda-corr, fig.height=7, fig.width=9}
cont_vars <- c(
  "kWh_total",
  "log_kWh_total",
  "temp_avg",
  "temp_range",
  "heating_degree_days",
  "cooling_degree_days",
  "humidity_avg",
  "precipitation",
  "sunshine_hours",
  "living_area",
  "kWh_per_m2"
)

cor_matrix <- heapo_model %>%
  select(all_of(cont_vars)) %>%
  cor(use = "pairwise.complete.obs") %>%
  round(2)

cor_long <- cor_matrix %>%
  as.data.frame() %>%
  tibble::rownames_to_column("Var1") %>%
  pivot_longer(-Var1, names_to = "Var2", values_to = "r")

ggplot(cor_long, aes(x = Var1, y = Var2, fill = r)) +
  geom_tile(color = "white") +
  geom_text(aes(label = r), size = 2.8) +
  scale_fill_gradient2(
    low = "tomato",
    mid = "white",
    high = "steelblue",
    midpoint = 0,
    limits = c(-1, 1),
    name = "r"
  ) +
  labs(
    title = "Pearson Correlation Matrix: Continuous Variables",
    x = NULL,
    y = NULL
  ) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 40, hjust = 1))
```

> **Interpretation:** The strongest correlations with `kWh_total` are:
>
> - **`heating_degree_days`** (strong positive): the most direct driver of consumption.
> - **`temp_avg`** (strong negative): colder days lead to more heating demand.
> - **`living_area`** (moderate positive): larger homes consume more.
>
> Note that `temp_avg` and `heating_degree_days` are highly correlated with each other, since HDD is derived from temperature (see HDD explanation above). Using both simultaneously in a model may cause multicollinearity; `heating_degree_days` is generally preferred as it directly quantifies heating demand rather than ambient temperature. `log_kWh_total` shows a slightly tighter linear relationship with temperature predictors than raw `kWh_total`, further justifying the log-transform.

---

## Binary Response: high_consumption

```{r eda-highcons, fig.height=4}
heapo_model %>%
  mutate(month_label = month(date, label = TRUE, abbr = TRUE)) %>%
  group_by(month_label) %>%
  summarise(pct_high = mean(high_consumption) * 100, .groups = "drop") %>%
  ggplot(aes(x = month_label, y = pct_high, fill = pct_high)) +
  geom_col(show.legend = FALSE) +
  geom_hline(yintercept = 50, linetype = "dashed", color = "grey50") +
  scale_fill_gradient(low = "lightblue", high = "tomato") +
  scale_y_continuous(limits = c(0, 100), labels = function(x) paste0(x, "%")) +
  labs(
    title = "% of Days with High Consumption (> overall median) by Month",
    subtitle = "Dashed line = 50% (expected if threshold were month-specific)",
    x = "Month",
    y = "% high consumption days"
  ) +
  theme_minimal()
```

> **Interpretation:** `high_consumption` is defined relative to the global median, so in winter months (when consumption is generally high) a much larger proportion of days are classified as high-consumption (up to roughly 90% in January). In summer, fewer than 10% of days exceed the global median. This strong class imbalance within seasons is important for GLM Binomial: the model needs to capture seasonal patterns rather than predicting a near-constant value.

---

# Summary

```{r summary-table}
tibble(
  Item = c(
    "Final dataset",
    "Households",
    "Observations",
    "Time window",
    "Eligibility criterion",
    "Sampling",
    "Primary response (LM/GAM)",
    "Binary response (GLM Binomial)",
    "Count response (GLM Poisson)",
    "Key predictors"
  ),
  Value = c(
    "heapo_modelling.rds / .csv",
    as.character(n_distinct(heapo_model$Household_ID)),
    comma(nrow(heapo_model)),
    "2022-01-01 to 2024-02-29",
    ">=95% data coverage in window",
    paste0(
      "Random sample of ",
      length(sampled_ids),
      " from ",
      nrow(eligible),
      " eligible (seed = 42, max 100k rows)"
    ),
    "log(kWh_total)",
    "high_consumption (kWh_total > global median)",
    "n_residents",
    "heating_degree_days, temp_avg, living_area, building_type, heatpump_type"
  )
) %>%
  styled_table("Final modelling dataset: summary")
```

---

*Report generated on `r Sys.Date()` using R `r R.version$major`.`r R.version$minor`.*
