---
title: "HEAPO – Linear Model"
author: ""
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float:
      collapsed: false
    code_folding: hide
    theme: flatly
    df_print: paged
    number_sections: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = FALSE,
  fig.width = 10,
  fig.height = 5
)
knitr::opts_knit$set(root.dir = normalizePath(".."))

library(here)
library(tidyverse)
library(broom)
library(car)
library(knitr)
library(kableExtra)

set.seed(42)

styled_table <- function(df, caption = "") {
  df %>%
    kable(caption = caption, digits = 4) %>%
    kable_styling(
      bootstrap_options = c("striped", "hover", "condensed"),
      full_width = FALSE
    )
}
```

```{r load-data, include=FALSE}
heapo_modelling <- readRDS(here(
  "data_processed",
  "heapo",
  "heapo_modelling.rds"
))
```

# Introduction

This section fits a Linear Model (LM) to predict daily household energy consumption. The response variable is `log_kWh_total` — the natural log of total daily kWh usage. Log-transforming energy amounts is standard practice: energy consumption is a strictly positive "amount" variable whose distribution is right-skewed on the original scale and approximately normal after log transformation.

The panel structure of the data (multiple daily observations per household) means the i.i.d. assumption of standard OLS is not fully met — standard errors may be slightly understated due to within-household correlation. Results should be interpreted with this caveat in mind.

# Univariate Models

Before fitting a full model, we compare each predictor individually to understand its marginal association with energy consumption.

```{r univariate-models}
predictors <- c(
  "heating_degree_days",
  "sunshine_hours",
  "humidity_avg",
  "living_area",
  "n_residents",
  "building_type",
  "heatpump_type"
)

model_comparison <- predictors %>%
  map_df(
    ~ {
      formula <- as.formula(paste("log_kWh_total ~", .x))
      lm(formula, data = heapo_modelling) %>%
        broom::glance() %>%
        mutate(predictor = .x)
    }
  ) %>%
  select(predictor, r.squared, adj.r.squared, p.value) %>%
  arrange(desc(r.squared))

styled_table(model_comparison, caption = "Univariate model R² by predictor")
```

`heating_degree_days` (cold-weather demand) is expected to be the strongest predictor of daily consumption. The table above confirms which predictors carry the most explanatory power individually.

# Simple Multiple Linear Regression

We first fit a model with all seven predictors and no interactions. This serves as our interpretable baseline.

```{r simple-mlr}
heapo.mlm <- lm(
  log_kWh_total ~ heating_degree_days +
    sunshine_hours +
    humidity_avg +
    living_area +
    n_residents +
    building_type +
    heatpump_type,
  data = heapo_modelling
)

summary(heapo.mlm)
```

**Interpretation of key coefficients:**

- `heating_degree_days`: Each additional heating degree day is associated with a positive increase in log energy consumption — colder days drive higher usage.
- `sunshine_hours`: More sunshine is associated with lower consumption, likely reflecting lower heating demand on bright days.
- `living_area`: Larger homes consume more energy, as expected.
- `n_residents`: More occupants increase energy demand.
- `building_type` / `heatpump_type`: Categorical effects capture structural and system-level differences in baseline consumption.

All coefficients are on the log scale: a coefficient of 0.05 means a 5% multiplicative change in kWh for a one-unit increase in the predictor.

# Extended Model with Interactions

A common pitfall when comparing LMs to non-parametric methods is using an overly simple LM. To ensure a fair comparison, we add two interaction terms:

- `heating_degree_days × heatpump_type`: different heat pump technologies (e.g., air-source vs. ground-source) have different efficiencies at cold temperatures, so the slope of consumption vs. cold should vary by pump type.
- `heating_degree_days × building_type`: well-insulated buildings are less sensitive to outdoor temperature than older, poorly-insulated ones.

```{r extended-mlr}
heapo.mlm.ext <- lm(
  log_kWh_total ~ heating_degree_days +
    sunshine_hours +
    humidity_avg +
    living_area +
    n_residents +
    building_type +
    heatpump_type +
    heating_degree_days:heatpump_type +
    heating_degree_days:building_type,
  data = heapo_modelling
)

summary(heapo.mlm.ext)
```

The interaction coefficients tell us whether the effect of cold weather on consumption differs by pump type or building type. A significant positive interaction means that group responds more strongly to cold; a negative interaction means it is better insulated against temperature swings.

# Model Diagnostics

We check the four standard LM assumptions using the built-in diagnostic plots.

```{r diagnostics, fig.height=8}
par(mfrow = c(2, 2))
plot(heapo.mlm.ext, main = "Extended MLR Diagnostics")
par(mfrow = c(1, 1))
```

- **Residuals vs Fitted**: Checks linearity. A horizontal band with no pattern is ideal.
- **Q-Q Plot**: Checks normality of residuals. Points near the diagonal indicate normality.
- **Scale-Location**: Checks homoscedasticity (constant variance). A flat line is ideal.
- **Residuals vs Leverage**: Identifies influential observations.

```{r vif}
vif_vals <- vif(heapo.mlm.ext)
vif_df <- data.frame(
  Predictor = names(vif_vals),
  VIF = round(as.numeric(vif_vals), 2)
)
styled_table(
  vif_df,
  caption = "Variance Inflation Factors (VIF) — extended model"
)
```

VIF values above 10 indicate problematic multicollinearity. `heating_degree_days` will naturally have elevated VIF due to its interaction terms — this is expected and does not invalidate the model.

# Coefficient Interpretation

The plot below shows the estimated coefficients of the extended model with 95% confidence intervals. Coefficients whose intervals do not cross zero are statistically significant.

```{r coef-plot, fig.height=7}
cis <- confint(heapo.mlm.ext)

coef_df <- tibble(
  term = names(coef(heapo.mlm.ext)),
  estimate = coef(heapo.mlm.ext),
  lower = cis[, 1],
  upper = cis[, 2]
) %>%
  filter(term != "(Intercept)")

ggplot(coef_df, aes(x = estimate, y = reorder(term, estimate))) +
  geom_point(size = 2, color = "#2c3e50") +
  geom_errorbarh(
    aes(xmin = lower, xmax = upper),
    height = 0.3,
    color = "#2c3e50"
  ) +
  geom_vline(xintercept = 0, linetype = "dashed", color = "#e74c3c") +
  theme_minimal(base_size = 13) +
  labs(
    title = "Extended MLR: Coefficient Estimates with 95% CIs",
    subtitle = "Coefficients on log scale — positive = higher consumption",
    x = "Coefficient estimate",
    y = NULL
  )
```

# Model Comparison

We compare the simple and extended models on adjusted R² and in-sample RMSE.

```{r model-comparison}
rmse <- function(model) sqrt(mean(residuals(model)^2))

comparison <- tibble(
  Model = c("Simple MLR (main effects)", "Extended MLR (+ interactions)"),
  `Adj. R²` = c(
    round(summary(heapo.mlm)$adj.r.squared, 4),
    round(summary(heapo.mlm.ext)$adj.r.squared, 4)
  ),
  RMSE = c(
    round(rmse(heapo.mlm), 4),
    round(rmse(heapo.mlm.ext), 4)
  )
)

styled_table(comparison, caption = "LM model comparison")
```

The extended model is expected to show a modest improvement in adjusted R² due to the interaction terms capturing heterogeneity across building and pump types. The improvement is intentionally kept focused — we avoid adding interactions for every predictor to retain interpretability.

# Save Model

```{r save-model}
saveRDS(heapo.mlm.ext, here("models", "linear_model", "heapo_lm.rds"))
cat("Model saved to models/linear_model/heapo_lm.rds\n")
```
