---
title: "HEAPO – Linear Model"
author: ""
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float:
      collapsed: false
    code_folding: hide
    theme: flatly
    df_print: paged
    number_sections: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = FALSE,
  fig.width = 10,
  fig.height = 5
)
knitr::opts_knit$set(root.dir = normalizePath(".."))

library(here)
library(tidyverse)
library(broom)
library(car)
library(knitr)
library(kableExtra)

set.seed(42)

styled_table <- function(df, caption = "") {
  df %>%
    kable(caption = caption, digits = 4) %>%
    kable_styling(
      bootstrap_options = c("striped", "hover", "condensed"),
      full_width = FALSE
    )
}
```

```{r load-data, include=FALSE}
heapo_modelling <- readRDS(here("data_processed", "heapo", "heapo_modelling.rds"))
```

# Introduction

This section fits a Linear Model (LM) to predict daily household energy consumption. The response variable is `log_kWh_total` — the natural log of total daily kWh usage. Log-transforming energy amounts is standard practice: energy consumption is a strictly positive "amount" variable whose distribution is right-skewed on the original scale and approximately normal after log transformation.

The panel structure of the data (multiple daily observations per household) means the i.i.d. assumption of standard OLS is not fully met — standard errors may be slightly understated due to within-household correlation. Results should be interpreted with this caveat in mind.

# Univariate Models

Before fitting a full model, we compare each predictor individually to understand its marginal association with energy consumption.

```{r univariate-models}
predictors <- c(
  "heating_degree_days",
  "sunshine_hours",
  "humidity_avg",
  "living_area",
  "n_residents",
  "building_type",
  "heatpump_type"
)

model_comparison <- predictors %>%
  map_df(~ {
    formula <- as.formula(paste("log_kWh_total ~", .x))
    lm(formula, data = heapo_modelling) %>%
      broom::glance() %>%
      mutate(predictor = .x)
  }) %>%
  select(predictor, r.squared, adj.r.squared, p.value) %>%
  arrange(desc(r.squared))

styled_table(model_comparison, caption = "Univariate model R² by predictor")
```

`heating_degree_days` (cold-weather demand) is expected to be the strongest predictor of daily consumption. The table above confirms which predictors carry the most explanatory power individually.
