---
title: "HEAPO – Support Vector Machine Classification"
author: ""
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float:
      collapsed: false
    code_folding: hide
    theme: flatly
    df_print: paged
    number_sections: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = FALSE,
  fig.width = 10,
  fig.height = 5
)
knitr::opts_knit$set(root.dir = normalizePath(".."))

library(here)
library(tidyverse)
library(e1071)
library(caret)
library(knitr)
library(kableExtra)

set.seed(42)

styled_table <- function(df, caption = "") {
  kable(df, caption = caption, digits = 4) %>%
    kable_styling(
      bootstrap_options = c("striped", "hover", "condensed"),
      full_width = FALSE
    )
}
```

```{r load-data, include=FALSE}
heapo_modelling <- readRDS(here("data_processed", "heapo", "heapo_modelling.rds"))
```

# Introduction

This section fits a Support Vector Machine (SVM) to predict daily household energy consumption. The response variable is `log_kWh_total` — the natural log of total daily kWh usage. SVMs are powerful non-parametric methods that can capture complex relationships through kernel transformations while remaining computationally efficient.

The panel structure of the data (multiple daily observations per household) means standard assumptions may not be fully met. Results should be interpreted with this caveat in mind.

