---
title: "HEAPO – Support Vector Machine Classification"
author: ""
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float:
      collapsed: false
    code_folding: hide
    theme: flatly
    df_print: paged
    number_sections: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = FALSE,
  fig.width = 10,
  fig.height = 5
)
knitr::opts_knit$set(root.dir = normalizePath(".."))

library(here)
library(tidyverse)
library(e1071)
library(caret)
library(knitr)
library(kableExtra)

set.seed(42)

styled_table <- function(df, caption = "") {
  df %>%
    kable(caption = caption, digits = 4) %>%
    kable_styling(
      bootstrap_options = c("striped", "hover", "condensed"),
      full_width = FALSE
    )
}
```

```{r load-data, include=FALSE}
heapo_modelling <- readRDS(here("data_processed", "heapo", "heapo_modelling.rds"))
```

# Introduction

Support Vector Machines (SVMs) are powerful classifiers that find the optimal hyperplane separating two classes by maximising the margin between the nearest data points (support vectors). Unlike the Linear Model, SVMs are non-parametric and can capture complex, non-linear decision boundaries via kernel functions.

**Classification task:** We classify each day as a *high-consumption day* (top 25% of daily kWh usage) or a *normal day*. This binary framing is practically useful: identifying high-load days in advance allows grid operators and energy providers to plan capacity and issue consumption alerts.

The top-quartile threshold is data-driven: it identifies the 25% of days with the highest energy demand, which are disproportionately responsible for grid stress events.

# Data Preparation

## Create Binary Target

```{r create-target}
q75 <- quantile(heapo_modelling$log_kWh_total, 0.75)
cat("Q75 threshold (log kWh):", round(q75, 3), "\n")
cat("Equivalent kWh:", round(exp(q75), 1), "\n")

heapo_svm_data <- heapo_modelling %>%
  mutate(
    high_consumption = as.factor(
      ifelse(log_kWh_total >= q75, "high", "normal")
    )
  ) %>%
  select(
    high_consumption,
    heating_degree_days,
    sunshine_hours,
    humidity_avg,
    living_area,
    n_residents,
    building_type,
    heatpump_type
  ) %>%
  na.omit()

table(heapo_svm_data$high_consumption)
```

The binary target `high_consumption` is excluded from the predictors. Note that `log_kWh_total` is also excluded to avoid data leakage — the SVM must predict high-load days from weather and household characteristics alone.

## Train / Test Split

```{r train-test-split}
set.seed(42)
train_idx <- createDataPartition(
  heapo_svm_data$high_consumption,
  p    = 0.8,
  list = FALSE
)

train_data <- heapo_svm_data[train_idx, ]
test_data  <- heapo_svm_data[-train_idx, ]
test_x     <- test_data %>% select(-high_consumption)
test_y     <- test_data$high_consumption

cat("Training rows:", nrow(train_data), "\n")
cat("Test rows:    ", nrow(test_data), "\n")
cat("Class balance (train):\n")
print(prop.table(table(train_data$high_consumption)))
```

An 80/20 stratified split preserves the class ratio in both sets.

# Baseline SVM (Radial Kernel, cost = 1)

We start with a radial basis function (RBF) kernel and default cost = 1. The RBF kernel maps data into a high-dimensional space, allowing non-linear decision boundaries — well-suited to energy consumption patterns that interact across weather and household variables.

`scale = TRUE` ensures all numeric features are standardised before training, which is essential for SVMs (they are sensitive to feature scale).

```{r svm-baseline, cache=TRUE}
set.seed(42)
svm_baseline <- svm(
  high_consumption ~ .,
  data   = train_data,
  kernel = "radial",
  cost   = 1,
  scale  = TRUE
)
summary(svm_baseline)
```

```{r svm-baseline-eval}
pred_baseline <- predict(svm_baseline, test_x)
cm_baseline   <- confusionMatrix(pred_baseline, test_y, positive = "high")
cm_baseline
```

# Tuned SVM (Cross-Validated)

We use 5-fold cross-validation to search over a small grid of `cost` and `gamma` values. `cost` controls the penalty for misclassification (higher = tighter fit), while `gamma` controls the influence radius of each training point (higher = more local boundaries).

```{r svm-tune, cache=TRUE}
set.seed(42)
svm_tune <- tune(
  svm,
  high_consumption ~ .,
  data          = train_data,
  kernel        = "radial",
  ranges        = list(
    cost  = c(1, 10, 100),
    gamma = c(0.01, 0.1)
  ),
  tunecontrol   = tune.control(cross = 5)
)

# Best parameters
svm_tune$best.parameters
```

```{r svm-tune-results}
tune_results <- svm_tune$performances %>%
  arrange(error) %>%
  rename(CV_error = error)

styled_table(tune_results, caption = "CV tuning results — lower error is better")
```

```{r svm-best-eval}
best_svm  <- svm_tune$best.model
pred_best <- predict(best_svm, test_x)
cm_best   <- confusionMatrix(pred_best, test_y, positive = "high")
cm_best
```
